---
title: "HMPV Co-detections Comprehensive Analysis"
author: "Ray Pomponio"
date: today
output-file: "Comprehensive Analysis Apr 2026.html"
format:
  html:
    embed-resources: true
    df-print: paged
execute:
  message: false
  warning: false
fig-cap-location: top
---

# Analytical Workflow

## Setup

Loading various `R` packages and setting plotting parameters...

```{r}
library(data.table)
library(consort)
library(grid)
library(gtsummary)
library(ggplot2)
library(ggpubr)
library(ggupset)
set_consort_defaults(txt_gp=gpar(cex=0.6), parse_markup=TRUE)
theme_set(theme_pubclean(base_size=14))
```

The following code reads the latest dataset as of April 2026:

```{r}
# load data from CDC; perform clean-up on dates and sites
cdc <- fread("Data/Pitt_Anna_HMPV_APR26.csv")
cdc[, c_ariyear:=factor(
  c_ariyear,
  c(1:10),
  c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20", "2020-21",
    "2021-22", "2022-23", "2023-24", "2024-25"))]
cdc[, studysite:=factor(
  studysite,
  c(1:6, 8),
  c("Vanderbilt", "Rochester", "Cincinnati",  "Seattle", "Houston",
    "Kansas City", "Pittsburgh"))]
dat <- cdc[research_tested==1 & !is.na(c_finalstatus)]
dat[is.na(respvisitdt), respvisitdt:=as.IDate(as.Date(scrdate, format="%m/%d/%Y"))]
```

## Inclusion/Exclusion Criteria

The only inclusion criteria is to have available research testing and determinable "final status," based on a 10-day chart review. Incidentally, not all cases were tested using the research panel. The following CONSORT figure describes inclusion/exclusion into this analysis, by site:

```{r}
#| label: fig-consort-plot1
#| fig-cap: "Study inclusion diagram"
#| echo: false
cdc[is.na(c_finalstatus), rem:="Final status undetermined"]
cdc[research_tested!=1, exc:="Sample not tested"]
out <- consort_plot(
  cdc,
  order=c(Caseid="CDC Case Count", rem="Invalid Case ID",
          studysite="Site Enrollments",
          exc="**Excluded**", Caseid="Analyzable Cases"),
  side_box=c("rem", "exc"),
  allocation="studysite")
plot(out, grViz=TRUE)
```

## Derivation of Outcome

Similar to our work in mBio (2026), we derive an ordinal outcome based on observable clinical events. For each outcome after Routine/ED discharge, events indicate escalating thresholds of "illess severity":

1) Routine/ED Discharge
2) Admitted to Hospital
3) Admitted, Received Supplemental O2 (Nasal Canula, HFNC, CPAP/BIPAP)
4) Admitted to ICU
5) Intubated or Treated with Mechanical Ventilation or ECMO
6) Died

```{r}
dat[, c_illsevnum:=fcase(
  c_died==1, 6,
  c_intubated==1, 5,
  inptEcmo==1, 5,
  inptICU==1, 4,
  c_finalstatus==1 & c_respsup==1, 3,
  c_finalstatus==1 & c_suppoxy==1, 3,
  c_finalstatus==1 & c_blowby==1, 3,
  c_finalstatus==1 & c_hfnc==1, 3,
  c_finalstatus==1 & c_cpap==1, 3,
  c_finalstatus==1, 2,
  c_finalstatus>=2, 1)]
dat[, c_severity:=factor(c_illsevnum, 1:6, c("1) Routine/ED Discharge",
                                             "2) Admitted to Hospital",
                                             "3) Admitted + SO2",
                                             "4) Admitted to ICU",
                                             "5) MV/ECMO", "6) Died"))]
```

The following plot demonstrates the overall distribution of severity levels over the period of enrollments:

```{r}
#| label: fig-outcome-seasonal
#| fig-cap: "Distribution of ordinal outcome by season"
#| echo: false
severity.counts <- dat[, .N, by=.(c_ariyear, c_severity)]
severity.counts[, perc:=N / sum(N), by=c_ariyear]
ggplot(severity.counts, aes(x=as.integer(c_ariyear), y=perc, fill=c_severity)) +
  geom_area(alpha=0.8, color="black", size=0.3) +
  scale_fill_brewer(palette=8) +
  geom_text(aes(label=format(100 * perc, digits=1)),
            position=position_fill(0.5),
            alpha=0.8, color="black", size=3.0) +
  # scale_y_reverse(breaks=seq(0, 1, 0.25), labels=seq(100, 0, -25)) +
  scale_y_continuous(breaks=seq(0, 1, 0.25), labels=seq(0, 100, 25)) +
  scale_x_continuous(breaks=1:10,
                     labels=levels(severity.counts$c_ariyear)) +
  labs(x="Season", y="Marginal Percentage", fill="Ordinal Illness Severity")
```

Although our interest is in severity as captured by observable events, it is possible to demonstrate the relationship between our derived ordinal outcome and the more-traditional Length of Stay (LOS) outcome, which might be treated continuously and normally distributed:

```{r}
#| label: fig-outcome-los-comp
#| fig-cap: "Relationship between ordinal outcome and log-transformed length of stay"
#| echo: false
dat[!is.na(c_los), c_los0:=c_los]
dat[c_finalstatus>=2, c_los0:=0]
ggplot(dat, aes(x=c_illsevnum, y=c_los0, color=c_severity)) +
  geom_point(position=position_jitter(width=0.1, seed=1818)) +
  geom_boxplot(alpha=0.8, outliers=FALSE) +
  scale_color_brewer(palette=8) +
  scale_x_continuous(breaks=1:6, labels=paste("Level", 1:6)) +
  scale_y_continuous(transform="pseudo_log") +
  labs(x="Ordinal Illness Severity", y="Length of Stay (days)", color="Legend")
```

@fig-outcome-los-comp makes clear the asssumption of normally distributed lengths of stay is violated; it is arguable whether the variable is continuous or discrete. More concerning, the most-severe outcome (death) does not yield the highest typical length of stay. As a fix for these numerical issues, any use of LOS as an outcome should consider, at the very least, a log-transformation.

## Derivation of Exposure

We next derive a set of exposures based on laboratory-confirmed detections; these exposures are the viral pathogens, in which any combination is theoretically possible. 

```{r}
dat[, c_mpv_result:=fcase(tmpv==1, 1, default=0)]
dat[, c_rsv_result:=fcase(c_rsv_result==1, 1, default=0)]
dat[, c_adeno_result:=fcase(tAdeno==1, 1, default=0)]
dat[, c_flu_result:=fcase(anyflu_result==1, 1, default=0)]
dat[, c_piv_result:=fcase(piv14_pos==1, 1, default=0)]
dat[, c_rhino_result:=fcase(rhent_pos==1, 1, default=0)]
lab.paths <- c(MPV="c_mpv_result", RSV="c_rsv_result", Adeno="c_adeno_result",
               Flu="c_flu_result", PIV="c_piv_result", Rhino="c_rhino_result")
custom_fun <- function(results) {
  # returns list of pathogens detected
  names(lab.paths)[results==1]
}
dat[, lst_detections:=apply(.SD, 1, custom_fun), .SDcols=unname(lab.paths)]
```

We examine the most common co-detections involving human metapneumovirus:

```{r}
#| label: fig-exposure-upset
#| fig-cap: "Most common co-detection patterns involving HMPV-positive cases"
#| echo: false
ggplot(dat[tmpv==1], aes(x=lst_detections)) +
  geom_bar() +
  scale_x_upset(n_intersections=15) +
  geom_text(stat="count", aes(label=after_stat(count)), vjust=-1) +
  scale_y_continuous(breaks=NULL, lim=c(0, 5000), name="") +
  labs(x="Co-detections with HMPV")
```

