---
title: "HMPV Illness Severity and Co-detections: NVSN 2015–2026"
subtitle: "Replication of Amarin et al. (2026) RSV Framework Applied to HMPV"
author: "Ray Pomponio / Anna Wang-Erickson"
date: today
output-file: "Replication Analysis May 2026.html"
format:
  html:
    embed-resources: true
    df-print: paged
    toc: true
    toc-depth: 3
execute:
  message: false
  warning: false
fig-cap-location: top
---

# Overview

This document replicates, as closely as the NVSN dataset permits, the analytical
approach of Amarin et al. (2026) for RSV, applied here to **human metapneumovirus
(HMPV)** as the pathogen of interest. The target outputs are:

- A CONSORT-style inclusion/exclusion diagram (by site)  
- Derivation and distribution of the ordinal illness-severity outcome  
- LOS validation of the ordinal outcome  
- Co-detection patterns (UpSet plot)  
- **Table 1**: Demographic and clinical characteristics of HMPV-positive cases,
  stratified by co-detection status  
- **Table 2**: Illness-severity distribution stratified by co-detection status  
- **Figure**: Seasonal distribution of HMPV cases and illness severity over time  

---

# Setup

```{r setup}
library(data.table)
library(consort)
library(grid)
library(gtsummary)
library(ggplot2)
library(ggpubr)
library(ggupset)

set_consort_defaults(txt_gp = gpar(cex = 0.6), parse_markup = TRUE)
theme_set(theme_pubclean(base_size = 14))
```

---

# Data Loading and Cleaning

```{r load-data}
# ── Load CSV (update path as needed) ──────────────────────────────────────────
cdc <- fread("Data/Pitt_Anna_HMPV_APR26.csv")

# ── ARI season labels (11 seasons, 2015-16 through 2025-26) ──────────────────
cdc[, c_ariyear := factor(
  c_ariyear,
  levels = 1:11,
  labels = c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20",
             "2020-21", "2021-22", "2022-23", "2023-24", "2024-25", "2025-26")
)]

# ── Site labels ───────────────────────────────────────────────────────────────
cdc[, studysite := factor(
  studysite,
  levels = c(1:6, 8),
  labels = c("Vanderbilt", "Rochester", "Cincinnati", "Seattle",
             "Houston", "Kansas City", "Pittsburgh")
)]

# ── Sex labels ────────────────────────────────────────────────────────────────
cdc[, sexch := factor(sexch, levels = c(1, 2, 8),
                      labels = c("Male", "Female", "Unknown"))]

# ── Insurance labels ──────────────────────────────────────────────────────────
cdc[, scrinsurance := factor(
  scrinsurance,
  levels = c(1, 2, 3, 4, 8),
  labels = c("Public", "Private", "Both", "None/Self-Pay", "Unknown")
)]

# ── Race/ethnicity (use c_race_int2 for finer granularity) ───────────────────
cdc[, c_race_int2 := factor(
  c_race_int2,
  levels = 1:9,
  labels = c("White NH", "Black NH", "Hispanic", "NH Pacific Islander",
             "AI/AN NH", "Asian NH", "Multiple/Other NH", "Unknown", "Unknown")
)]

# ── Coerce binary flags to integer 0/1 (some may arrive as character) ─────────
binary_vars <- c("c_xunderlying", "c_uresp", "c_unuerologic", "c_uoncolimmu",
                 "c_ucardiovasc", "c_premature",
                 "c_respsup", "c_suppoxy", "inptICU", "inptEcmo",
                 "c_blowby", "c_hfnc", "c_cpap", "c_intubated", "c_died",
                 "ANYpneumDx", "ANYbronchDx", "ANYasthDx", "ANYuriDx",
                 "ANYrespfailDx")
cdc[, (binary_vars) := lapply(.SD, function(x) as.integer(as.character(x))),
    .SDcols = binary_vars]

# ── Backfill specimen date from screen date where missing ─────────────────────
cdc[is.na(respvisitdt),
    respvisitdt := as.IDate(as.Date(scrdate, format = "%m/%d/%Y"))]

# ── Analytic cohort: research-tested with determinable final status ────────────
dat <- cdc[research_tested == 1 & !is.na(c_finalstatus)]
```

---

# Inclusion / Exclusion (CONSORT Diagram)

```{r}
#| label: fig-consort
#| fig-cap: "Study inclusion diagram, by enrollment site"
#| echo: false

cdc[is.na(c_finalstatus), rem := "Final status undetermined"]
cdc[research_tested != 1,  exc := "Sample not research-tested"]

out <- consort_plot(
  cdc,
  order = c(Caseid    = "CDC Case Count",
            rem       = "Invalid / Missing Status",
            studysite = "Site Enrollments",
            exc       = "**Excluded**",
            Caseid    = "Analyzable Cases"),
  side_box   = c("rem", "exc"),
  allocation = "studysite"
)
plot(out, grViz = TRUE)
```


---

# Outcome Derivation: Ordinal Illness Severity

Following the mBio (2026) framework, we construct a six-level ordinal outcome
based on escalating observable clinical events. For admitted patients, the
**highest** event reached during the visit determines severity.

```{r derive-severity}
dat[, c_illsevnum := fcase(
  c_died      == 1, 6L,
  c_intubated == 1, 5L,
  inptEcmo    == 1, 5L,
  inptICU     == 1, 4L,
  c_finalstatus == 1 & c_respsup == 1, 3L,
  c_finalstatus == 1 & c_suppoxy == 1, 3L,
  c_finalstatus == 1 & c_blowby  == 1, 3L,
  c_finalstatus == 1 & c_hfnc    == 1, 3L,
  c_finalstatus == 1 & c_cpap    == 1, 3L,
  c_finalstatus == 1,               2L,
  c_finalstatus >= 2,               1L   # ED / OP / UC
)]

sev_labels <- c("1) Routine/ED Discharge",
                "2) Admitted to Hospital",
                "3) Admitted + Supplemental O2",
                "4) Admitted to ICU",
                "5) MV / ECMO",
                "6) Died")

dat[, c_severity := factor(c_illsevnum, levels = 1:6, labels = sev_labels)]
```

## Seasonal Distribution of Severity

```{r}
#| label: fig-severity-seasonal
#| fig-cap: "Distribution of ordinal illness severity by ARI season (all cases with final status)"
#| echo: false

sev_counts <- dat[, .N, by = .(c_ariyear, c_severity)]
sev_counts[, perc := N / sum(N), by = c_ariyear]

ggplot(sev_counts,
       aes(x = as.integer(c_ariyear), y = perc, fill = c_severity)) +
  geom_area(alpha = 0.8, color = "black", linewidth = 0.3) +
  geom_text(aes(label = sprintf("%.0f", 100 * perc)),
            position = position_fill(0.5),
            alpha = 0.8, color = "black", size = 2.8) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  scale_y_continuous(breaks = seq(0, 1, 0.25),
                     labels = seq(0, 100, 25),
                     name  = "Marginal Percentage (%)") +
  scale_x_continuous(breaks  = 1:11,
                     labels  = levels(sev_counts$c_ariyear),
                     name    = "ARI Season") +
  labs(fill = "Ordinal Illness Severity") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

## Ordinal Severity vs. Length of Stay

```{r}
#| label: fig-severity-los
#| fig-cap: "Ordinal severity vs. log-transformed length of stay (inpatients)"
#| echo: false

# Assign LOS = 0 for non-admitted; keep NA for admitted with missing LOS
dat[!is.na(c_los), c_los0 := c_los]
dat[c_finalstatus >= 2, c_los0 := 0]

ggplot(dat, aes(x = c_illsevnum, y = c_los0, color = c_severity)) +
  geom_point(position = position_jitter(width = 0.15, seed = 1818),
             alpha = 0.4, size = 0.8) +
  geom_boxplot(alpha = 0.7, outliers = FALSE, linewidth = 0.7) +
  scale_color_brewer(palette = "RdYlBu", direction = -1) +
  scale_x_continuous(breaks = 1:6, labels = paste("Level", 1:6)) +
  scale_y_continuous(transform = "pseudo_log",
                     name = "Length of Stay (days, pseudo-log scale)") +
  labs(x = "Ordinal Illness Severity", color = "Legend") +
  theme(legend.position = "right")
```

> **Note (mirroring Amarin et al.):** LOS is non-normally distributed and does
> not monotonically increase with severity — notably, deaths (Level 6) do not
> produce the longest stays. Any secondary LOS analysis should apply at minimum a
> log-transformation; ordinal severity remains the primary outcome.

---

# Exposure Derivation: Viral Detections

We derive binary detection flags for each pathogen and build a list-column of
all detected pathogens per case, enabling UpSet-plot visualization of
co-detection combinations.

```{r derive-exposures}
# ── Binary detection flags (research lab; treat inconclusive as negative) ──────
dat[, c_mpv_result   := fcase(tmpv          == 1, 1L, default = 0L)]
dat[, c_rsv_result2  := fcase(c_rsv_result  == 1, 1L, default = 0L)]
dat[, c_adeno_result := fcase(tAdeno        == 1, 1L, default = 0L)]
dat[, c_flu_result   := fcase(anyflu_result == 1, 1L, default = 0L)]
dat[, c_piv_result   := fcase(piv14_pos     == 1, 1L, default = 0L)]
dat[, c_rhino_result := fcase(rhent_pos     == 1, 1L, default = 0L)]

lab.paths <- c(
  HMPV  = "c_mpv_result",
  RSV   = "c_rsv_result2",
  Adeno = "c_adeno_result",
  Flu   = "c_flu_result",
  PIV   = "c_piv_result",
  Rhino = "c_rhino_result"
)

# ── List-column of detected pathogen names ─────────────────────────────────────
custom_fun <- function(results) names(lab.paths)[results == 1]
dat[, lst_detections := apply(.SD, 1, custom_fun), .SDcols = unname(lab.paths)]

# ── Co-detection flag (any co-pathogen among HMPV-positives) ──────────────────
dat[tmpv == 1,
    codetect_any := fcase(
      c_rsv_result2 == 1 | c_adeno_result == 1 |
        c_flu_result == 1 | c_piv_result  == 1 |
        c_rhino_result == 1, "Co-detection",
      default = "HMPV Only"
    )]
dat[tmpv == 1, codetect_any := factor(codetect_any,
                                      levels = c("HMPV Only", "Co-detection"))]
```

## Co-detection Patterns (UpSet Plot)

```{r}
#| label: fig-upset-hmpv
#| fig-cap: "Most common co-detection patterns among HMPV-positive cases (top 15 combinations)"
#| echo: false

ggplot(dat[tmpv == 1], aes(x = lst_detections)) +
  geom_bar(fill = "steelblue") +
  scale_x_upset(n_intersections = 15) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5,
            size = 3) +
  scale_y_continuous(name = "Number of Cases",
                     expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Co-detections with HMPV",
       caption = "Restricted to research-tested, HMPV-positive cases with determinable final status.")
```

---

# Table 1: Characteristics of HMPV-Positive Cases

Stratified by **any co-detection** (HMPV alone vs. HMPV + ≥1 co-pathogen).
Modeled after Table 1 of Amarin et al. (2026).

```{r}
#| label: tbl-table1
#| tbl-cap: "Demographic and clinical characteristics of HMPV-positive cases by co-detection status"

hmpv_pos <- dat[tmpv == 1]

tbl1_vars <- c(
  # Demographics
  "c_agemonths", "c_ageyears", "sexch", "c_race_int2", "scrinsurance",
  # Comorbidities
  "c_xunderlying", "c_uresp", "c_ucardiovasc", "c_unuerologic",
  "c_uoncolimmu", "c_premature",
  # Discharge diagnoses
  "ANYpneumDx", "ANYbronchDx", "ANYasthDx", "ANYuriDx", "ANYrespfailDx",
  # Severity
  "c_severity", "c_los"
)

# Only keep vars that exist in the data
tbl1_vars <- intersect(tbl1_vars, names(hmpv_pos))

hmpv_pos[, (tbl1_vars) := lapply(.SD, function(x) {
  if (is.integer(x) && all(x %in% c(0L, 1L, NA_integer_))) factor(x, 0:1, c("No","Yes"))
  else x
}), .SDcols = tbl1_vars]

tbl1 <- hmpv_pos |>
  tbl_summary(
    by       = codetect_any,
    include  = all_of(tbl1_vars),
    label    = list(
      c_agemonths     ~ "Age (months)",
      c_ageyears      ~ "Age (years)",
      sexch           ~ "Sex",
      c_race_int2     ~ "Race / Ethnicity",
      scrinsurance    ~ "Insurance Type",
      c_xunderlying   ~ "≥1 Underlying Condition",
      c_uresp         ~ "Respiratory Condition",
      c_ucardiovasc   ~ "Cardiovascular Condition",
      c_unuerologic   ~ "Neuromuscular Condition",
      c_uoncolimmu    ~ "Oncologic/Immunosuppressive Condition",
      c_premature     ~ "Prematurity (<37 wks)",
      ANYpneumDx      ~ "Pneumonia Dx",
      ANYbronchDx     ~ "Bronchiolitis Dx",
      ANYasthDx       ~ "Asthma Dx",
      ANYuriDx        ~ "Upper Respiratory Infection Dx",
      ANYrespfailDx   ~ "Respiratory Failure Dx",
      c_severity      ~ "Ordinal Illness Severity",
      c_los           ~ "Length of Stay (days)"
    ),
    missing  = "ifany",
    statistic = list(
      all_continuous()  ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    )
  ) |>
  add_overall() |>
  add_p() |>
  bold_labels() |>
  modify_caption("**Table 1.** Demographic and clinical characteristics of HMPV-positive cases by co-detection status")

tbl1
```


---

# Table 2: Illness Severity by Co-detection Status

```{r}
#| label: tbl-table2
#| tbl-cap: "Ordinal illness severity distribution by co-detection status"

tbl2 <- hmpv_pos |>
  tbl_summary(
    by      = codetect_any,
    include = c_severity,
    label   = list(c_severity ~ "Ordinal Illness Severity"),
    statistic = list(all_categorical() ~ "{n} ({p}%)")
  ) |>
  add_overall() |>
  add_p() |>
  bold_labels() |>
  modify_caption("**Table 2.** Distribution of ordinal illness severity by co-detection status")

tbl2
```

---

# Seasonal HMPV Incidence

Mirroring Amarin et al. Figure 1: weekly HMPV case counts over the study period,
colored by illness severity.

```{r}
#| label: fig-hmpv-weekly
#| fig-cap: "Weekly HMPV-positive case counts by season, colored by ordinal illness severity"
#| echo: false

hmpv_weekly <- dat[tmpv == 1 & !is.na(respvisitdt) & !is.na(c_severity)]
hmpv_weekly[, epiweek := as.integer(format(as.Date(respvisitdt), "%V"))]
hmpv_weekly[, epiyear := as.integer(format(as.Date(respvisitdt), "%G"))]
hmpv_weekly[, week_date := as.Date(
  paste(epiyear, epiweek, "1", sep = "-"), format = "%G-%V-%u"
)]

weekly_counts <- hmpv_weekly[, .N, by = .(week_date, c_ariyear, c_severity)]

ggplot(weekly_counts,
       aes(x = week_date, y = N, fill = c_severity)) +
  geom_col(width = 6) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(x = "Date", y = "Weekly HMPV Cases",
       fill  = "Ordinal Illness Severity",
       caption = "Bars represent weekly counts of research-tested, HMPV-positive cases.") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

---

# Supplemental: HMPV CT Value Distribution

The `tmpvCT` variable provides quantitative PCR cycle threshold values, which
inversely correlate with viral load. This is not part of Amarin et al. but is
a useful descriptive supplement.

```{r}
#| label: fig-ct-dist
#| fig-cap: "Distribution of HMPV CT values by ordinal illness severity"
#| echo: false

hmpv_ct <- dat[tmpv == 1 & !is.na(tmpvCT) & tmpvCT > 0]

ggplot(hmpv_ct, aes(x = c_severity, y = tmpvCT, color = c_severity)) +
  geom_boxplot(outliers = FALSE, linewidth = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.25, size = 0.7) +
  scale_color_brewer(palette = "RdYlBu", direction = -1) +
  scale_x_discrete(labels = function(x) gsub("^[0-9]+\\) ", "", x)) +
  labs(x = "Ordinal Illness Severity",
       y  = "HMPV CT Value (lower = higher viral load)",
       color = "Severity") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# Notes and Caveats

- **Scope**: This document replicates the descriptive and tabular framework of
  Amarin et al. (2026) only. Regression modeling, subgroup analyses, and
  journal-club extensions are intentionally excluded at this stage.

- **c_respsup / c_blowby / c_hfnc / c_cpap** are only available for
  `c_ariyear ≥ 8` (2022-23 onward). For earlier seasons, supplemental-O2
  classification in Severity Level 3 relies on `c_suppoxy` alone. This
  introduces a potential under-ascertainment of Level 3 in earlier seasons;
  consider flagging this in any manuscript methods section.

- **Race/ethnicity**: Two variables are available (`c_race_int` and
  `c_race_int2`). The finer-grained `c_race_int2` is used here. Confirm with
  PI whether this is the preferred variable.

- **HMPV season year 11 (2025-26)** is ongoing as of the data pull date
  (Apr 2026). Consider censoring or labeling partial-season data accordingly.

- **`tadeno` vs. `tAdeno`**: The data dictionary lists `tadeno` (lowercase).
  The template script uses `tAdeno` — verify case in the actual CSV header.

- **Co-detection flags**: `c_othercoin6` / `c_hothercoin6` (pre-derived HMPV
  co-detection indicators) exist in the data dictionary. These could be used
  as a consistency check against the manually derived `codetect_any` flag above.

---

*Generated with Quarto. Contact Ray Pomponio / Anna Wang-Erickson for questions.*
