---
title: "HMPV Co-detections in NVSN: 2015-2026"
subtitle: "Investigating Illness Severity Effects"
author: "Ray Pomponio / Anna Wang-Erickson"
date: today
output-file: "Comprehensive Analysis June 2026.html"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
    df-print: paged
    toc: true
    toc-depth: 3
execute:
  message: false
  warning: false
fig-cap-location: top
---

# Overview

This document examines the association between viral co-detection and illness
severity in children with **human metapneumovirus (HMPV)**, using data from
the New Vaccine Surveillance Network (NVSN) spanning the 2015–16 through
2025–26 surveillance seasons. The analytical framework is informed by Amarin
et al. (2026), who studied co-detection effects in children with RSV, though
the present analysis departs from their approach in several ways described
below.

**Reference:** Amarin JZ, Shaikha AA, Bunsow LI, et al. Respiratory Syncytial
Virus Co-Detection With Other Respiratory Viruses Is Not Significantly
Associated With Worse Clinical Outcomes Among Children Aged <2 Years.
*Open Forum Infectious Diseases*. 2026. <https://doi.org/10.1093/cid/ciaf194>.

The target outputs are:

- A flow diagram of study participants among ARI-enrolled medical encounters.
- A table of baseline characteristics and outcomes of HMPV-positive cases,
  stratified by co-detection status.
- An UpSet[^1] plot of the most common co-detection patterns among HMPV-positives.
- A proportional odds regression summary of adjusted odds ratios for elevated
  illness severity, grouped by co-detection status.
- Two sensitivity analyses: (1) restricted to patients without pre-existing
  conditions, and (2) with surveillance season added as a natural cubic spline,
  to assess robustness of the primary findings.

[^1]: UpSet plots visualize combinations of categorical attributes (here, which viruses were co-detected).

---

# Setup

The following packages are required to run this analysis. Each serves a
distinct purpose:

- **`data.table`**: fast data manipulation for large surveillance datasets
- **`MASS`**: provides `polr()`, which fits the proportional odds (ordinal
  logistic) regression model used in @tbl-table2, @tbl-table3, and @tbl-table4
- **`splines`**: provides `ns()` for natural cubic spline basis expansion of
  the season covariate, used in @tbl-table4
- **`consort`**: draws the participant flow diagram
- **`grid`**: low-level graphics support used internally by `consort`
- **`gtsummary`**: produces publication-ready summary tables
- **`gt`**: underlying table engine used by `gtsummary`; required here to
  control column widths in @tbl-table1
- **`ggplot2`** and **`ggpubr`**: general-purpose plotting
- **`ggupset`**: extends `ggplot2` to render UpSet plots

```{r setup}
library(data.table)
library(MASS)
library(splines)
library(consort)
library(grid)
library(gtsummary)
library(gt)
library(ggplot2)
library(ggpubr)
library(ggupset)
source("wald_function.R")
set_consort_defaults(txt_gp=gpar(cex=0.6), parse_markup=TRUE)
theme_set(theme_pubclean(base_size=14))
```

---

# Data Loading and Cleaning

The raw dataset is a CSV file extracted from the NVSN surveillance system.
Before any analysis, several housekeeping steps are required:

1. **Derive hospitalization status** — a simple binary flag (hospitalized vs.
   not) used as a grouping variable in @tbl-table1.
2. **Derive ordinal illness severity** — a six-level ordered outcome ranging
   from routine or ED discharge (Level 1) to death (Level 6). This is the
   primary outcome in @tbl-table2 through @tbl-table4, and captures gradations
   of severity that a binary hospitalization flag would obscure.
3. **Apply human-readable labels** to coded categorical variables (season,
   site, sex, insurance, race/ethnicity).
4. **Coerce binary (0/1) clinical flags** to integer type.
5. **Backfill missing specimen dates** from the screen date.

```{r load-data}
# ── Load CSV (update path as needed) ──────────────────────────────────────────
cdc <- fread("Data/Pitt_Anna_HMPV_APR26.csv")

# ── Derive binary hospitalization flag ────────────────────────────────────────
#    c_finalstatus == 1 means the patient was admitted as an inpatient;
#    all other values (ED, outpatient, urgent care) are grouped as "Not Hospitalized"
cdc[, hospital_status:=fcase(
  c_finalstatus==1, "Hospitalized",
  default="Not Hospitalized")]

# ── Derive ordinal illness severity (primary outcome) ─────────────────────────
#    Six levels, assigned in descending priority so each case is classified
#    at its highest observed level of care. Levels 5 and 6 are sparse but
#    retained to avoid information loss at the severe end of the spectrum.
#    Level 1: ED, outpatient, or urgent care discharge
#    Level 2: Admitted to hospital (no supplemental oxygen)
#    Level 3: Admitted + supplemental oxygen (any modality)
#    Level 4: ICU admission
#    Level 5: Mechanical ventilation or ECMO
#    Level 6: Death
cdc[, c_severity := fcase(
  c_died      == 1L,  6L,
  c_intubated == 1L,  5L,
  inptEcmo    == 1L,  5L,
  inptICU     == 1L,  4L,
  hospital_status == "Hospitalized" & (c_suppoxy == 1L | c_blowby == 1L |
    c_hfnc == 1L | c_cpap == 1L),  3L,
  hospital_status == "Hospitalized",  2L,
  default = 1L)]
cdc[, c_severity := factor(c_severity,
  levels  = 1:6,
  labels  = c("Discharge", "Hospitalized", "Hospitalized + O2",
              "ICU", "Ventilated / ECMO", "Died"),
  ordered = TRUE)]

# ── ARI season labels (11 seasons, 2015-16 through 2025-26) ──────────────────
cdc[, c_ariyear := factor(
  c_ariyear,
  levels = 1:11,
  labels = c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20",
             "2020-21", "2021-22", "2022-23", "2023-24", "2024-25", "2025-26")
)]

# ── Integer season index for spline (1 = 2015-16, 11 = 2025-26) ───────────────
#    ns() requires a numeric covariate; c_ariyear_int is kept separate from
#    the factor c_ariyear which is used for labeling only
cdc[, c_ariyear_int:=as.integer(c_ariyear)]

# ── Site labels ───────────────────────────────────────────────────────────────
#    Site code 7 is not used; Pittsburgh is coded as 8
cdc[, studysite := factor(
  studysite,
  levels = c(1:6, 8),
  labels = c("Vanderbilt", "Rochester", "Cincinnati", "Seattle",
             "Houston", "Kansas City", "Pittsburgh")
)]

# ── Sex labels ────────────────────────────────────────────────────────────────
cdc[, sexch := factor(sexch, levels=c(1, 2), labels=c("Male", "Female"))]

# ── Insurance labels ──────────────────────────────────────────────────────────
#    Code 8 (Unknown) is excluded; unknown cases appear as missing in tables
cdc[, scrinsurance := factor(
  scrinsurance,
  levels = c(1, 2, 3, 4),
  labels = c("Public", "Private", "Both", "None/Self-Pay"))]

# ── Race/ethnicity ────────────────────────────────────────────────────────────
#    Using c_race_int2 (7 categories). NH = Non-Hispanic; AI/AN = American
#    Indian / Alaska Native. Code 8 (Unknown) excluded; appears as missing.
cdc[, c_race_int2 := factor(
  c_race_int2,
  levels = 1:7,
  labels = c("White NH", "Black NH", "Hispanic", "NH Pacific Islander",
             "AI/AN NH", "Asian NH", "Multiple/Other NH"))]

# ── Coerce binary clinical flags to integer 0/1 ───────────────────────────────
#    These variables may import as character strings; explicit coercion avoids
#    errors in table summaries and models.
binary.vars <- c("c_xunderlying", "c_uresp", "c_unuerologic", "c_uoncolimmu",
                 "c_ucardiovasc", "c_premature",
                 "c_respsup", "c_suppoxy", "inptICU", "inptEcmo",
                 "c_blowby", "c_hfnc", "c_cpap", "c_intubated", "c_died",
                 "ANYpneumDx", "ANYbronchDx", "ANYasthDx", "ANYuriDx",
                 "ANYrespfailDx")
cdc[, (binary.vars) := lapply(.SD, function(x) as.integer(as.character(x))),
    .SDcols = binary.vars]

# ── Backfill specimen date from screen date where missing ─────────────────────
cdc[is.na(respvisitdt),
    respvisitdt := as.IDate(as.Date(scrdate, format="%m/%d/%Y"))]
```

---

# Inclusion and Exclusion of Participants

Not every enrolled encounter is eligible for this analysis. The flow diagram
below tracks cases from initial enrollment through to the final analytic cohort,
making the reasons for exclusion explicit and auditable.

There are three grounds for exclusion, applied in order:

1. **No research testing** — the specimen was not tested using the research
   multiplex panel, so HMPV status cannot be determined.
2. **HMPV not tested** — the research panel was run, but the HMPV assay was
   not performed (result code 8 = "not performed").
3. **Final status undetermined** — the 10-day chart review needed to classify
   the encounter's highest level of care could not be completed.

After excluding these cases, HMPV-negative and inconclusive results are set
aside (shown as "Undetected" in @fig-consort). Among confirmed HMPV-positive
cases, those with **more than one co-detected virus** are dropped to keep
co-detection groups mutually exclusive and directly comparable. The remaining
cases form the analytic cohort, split by whether any co-pathogen was detected
alongside HMPV.

```{r}
#| label: fig-consort
#| fig-cap: "Flow diagram of study participants from ARI enrollment through the final analytic cohort. Cases are split at the bottom by co-detection status (HMPV alone vs. HMPV with one co-detected virus). Cases with more than one co-detected virus are dropped to maintain mutually exclusive comparison groups."
#| fig-width: 8
#| fig-height: 6
#| echo: false

# ── Define exclusion reasons (applied before HMPV result is known) ────────────
cdc[, exc:=fcase(
  research_tested!=1, "Sample not research-tested",
  tmpv==8, "Sample not research-tested for HMPV",
  is.na(c_finalstatus), "Final status undetermined")]

# ── Among tested cases, flag those without a positive HMPV result ─────────────
cdc[, rem:=fcase(
  tmpv==0, "HMPV Negative",
  tmpv==2, "HMPV Inconclusive")]

# ── Among HMPV-positives, classify co-detection status ────────────────────────
cdc[, codetect:=fcase(
  c_othercoin6==1, "At least 1 other virus detected",
  default="No other virus was detected")]

# ── Count how many co-pathogens were detected (to identify multi-co-detections)
cdc[c_othercoin6==1, codetectn:=rowSums(.SD==1),
  .SDcols=c("tAdeno", "c_rsv_result", "anyflu_result", "piv14_pos", "rhent_pos")]
cdc[codetectn > 1, more:="More than 1 other virus detected"]

# ── Final analytic cohort: no exclusions, no ambiguous HMPV result,
#    and at most one co-detected virus ─────────────────────────────────────────
dat <- cdc[is.na(exc) & is.na(rem) & (c_othercoin6==0 | is.na(more))]

consort_plot(
  cdc,
  order = c(Caseid    = "Enrolled from outpatient, emergency, or inpatient",
            exc       = "**Excluded**",
            Caseid    = "Tested for human metapneumovirus",
            rem       = "**Undetected**",
            Caseid    = "Human metapneumovirus detected by research testing",
            codetect  = "",
            more      = "**Dropped**",
            Caseid    = "Analyzed"),
  side_box   = c("rem", "exc", "more"),
  allocation = "codetect")
```

---

# Characteristics of HMPV-Positive Cases

The table below describes the baseline characteristics and clinical outcomes
of all cases in the analytic cohort, stratified by co-detection status.
Each column represents a distinct viral detection group: cases in which HMPV
was the only virus detected, and cases in which HMPV was detected alongside
one specific co-pathogen (adenovirus, influenza, parainfluenza [PIV],
RSV, or rhinovirus/enterovirus [HRV]).[**Human Coronavirus (HCoV)** results were withheld from this dataset for unknown reasons. Data on **daycare attendance** and **smoking exposure**, which Amarin et al. included as baseline characteristics, were also withheld and could not be incorporated.]{.aside}

**How to read this table:** For continuous variables (e.g., age in months),
we report the median and interquartile range (IQR). For categorical variables
(e.g., sex, insurance type), we report counts and column percentages. Rows
labeled "Unknown" or missing indicate data that were not collected or could
not be determined. The clinical outcome variables (length of stay, ICU
admission, supplemental oxygen, intubation) appear at the bottom of the table
under *Final Status* and are only meaningful for hospitalized cases;
non-hospitalized cases contribute zeros or missing values to those rows.
The table is paginated (11 rows per page) and horizontally scrollable to
accommodate the six co-detection columns.

```{r}
#| label: tbl-table1
#| tbl-cap: "Baseline demographic characteristics and clinical outcomes of HMPV-positive cases, stratified by co-detection status. Co-detection groups are mutually exclusive (cases with more than one co-detected virus were excluded). Outcome variables (length of stay, ICU admission, supplemental oxygen, and intubation) are shown beneath Final Status and pertain only to hospitalized cases."

# ── Derive the co-detection status grouping variable ─────────────────────────
#    Priority is assigned in the order listed (Adeno first, HRV last);
#    multi-detection cases have already been excluded, so this is a simple lookup
dat[, codetect_status:=fcase(
  tAdeno==1, "HMPV/Adeno",
  anyflu_result==1, "HMPV/Flu",
  piv14_pos==1, "HMPV/PIV",
  c_rsv_result==1, "HMPV/RSV",
  rhent_pos==1, "HMPV/HRV",
  default="HMPV Only")]

# ── Variables to include in Table 1, in display order ────────────────────────
tbl1.vars <- c(
  # Demographics
  "c_agemonths", "sexch", "c_race_int2", "scrinsurance", "studysite",
  # Comorbidities
  "c_xunderlying", "c_uresp", "c_ucardiovasc", "c_unuerologic",
  "c_uoncolimmu", "c_premature",
  # Clinical outcomes (hospital_status first; inpatient-only outcomes follow)
  "hospital_status", "c_los", "inptICU", "c_suppoxy", "c_intubated")

tbl1 <- tbl_summary(
  dat,
  by       = codetect_status,
  include  = all_of(tbl1.vars),
  label    = list(
      c_agemonths     ~ "Age in months, median (IQR)",
      sexch           ~ "Sex",
      c_race_int2     ~ "Race / Ethnicity",
      scrinsurance    ~ "Insurance Type",
      studysite       ~ "Study Site",
      c_xunderlying   ~ "≥1 Underlying Condition",
      c_uresp         ~ "Respiratory Condition",
      c_ucardiovasc   ~ "Cardiovascular Condition",
      c_unuerologic   ~ "Neuromuscular Condition",
      c_uoncolimmu    ~ "Oncologic/Immunosuppressive Condition",
      c_premature     ~ "Prematurity (<37 wks)",
      hospital_status ~ "Final Status",
      c_los           ~ "Length of stay in days, median (IQR)",
      inptICU         ~ "ICU Admission",
      c_suppoxy       ~ "Supplemental Oxygen",
      c_intubated     ~ "Intubation"),
    missing  = "ifany",
    statistic = list(
      all_continuous()  ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)")) |>
  bold_labels() |>
  remove_footnote_header() |>
  modify_header(all_stat_cols() ~ "**{level}**") |>
  as_gt() |>
  cols_width(
    label        ~ px(220),
    everything() ~ px(95)) |>
  opt_interactive(use_compact_mode=TRUE, page_size_default=11)
tbl1
```

---

# Co-detection Patterns

The UpSet plot below visualizes how often HMPV was detected in isolation versus
each possible combination of the other five respiratory viruses tested in the
research panel (RSV, adenovirus, influenza, parainfluenza, and
rhinovirus/enterovirus). Each bar represents a unique combination of detected
viruses; the dots and connecting lines below the bars indicate which viruses
are included in that combination. Bars are colored by whether the patient was
ultimately hospitalized, providing a first visual impression of whether certain
co-detection patterns are associated with greater illness severity.

Note that this figure is drawn from the **HMPV-positive universe** (before
the restriction to single co-detections), so all combinations — including those
with two or more co-pathogens — are visible here. Only the top 15 most common
combinations are shown.

```{r}
#| label: fig-upset-hmpv
#| fig-cap: "Co-detection patterns among all HMPV-positive cases (research-tested), showing the 15 most frequent combinations of co-detected viruses. Bars are colored by hospitalization status. This figure is drawn before the restriction to single co-detections, so combinations involving multiple co-pathogens are included."
#| echo: false

# ── Named vector mapping display labels to raw column names ───────────────────
lab.paths <- c(
  HMPV  = "tmpv",
  RSV   = "c_rsv_result",
  Adeno = "tAdeno",
  Flu   = "anyflu_result",
  PIV   = "piv14_pos",
  Rhino = "rhent_pos")

# ── Build a list-column: for each case, which pathogens were detected? ─────────
#    customFun returns the names of detected pathogens (value == 1) per row;
#    stored as a list-column because the number of detections varies by case
customFun <- function(results) names(lab.paths)[results == 1]
cdc[, lst_detections:=apply(.SD, 1, customFun), .SDcols=unname(lab.paths)]

ggplot(cdc[tmpv==1], aes(x=lst_detections, fill=hospital_status)) +
  geom_bar(position=position_dodge(width=1)) +
  scale_x_upset(n_intersections=15) +
  geom_text(stat="count", aes(label=after_stat(count)),
            position=position_dodge(width=1), vjust=-0.5, size=3) +
  scale_y_continuous(name="Number of Cases",
                     expand=expansion(mult=c(0, 0.12))) +
  labs(x="Co-detections with HMPV", fill="Final Status")
```

---

# Illness Severity Odds by Co-detection Status

To assess whether the presence of a viral co-pathogen alongside HMPV is
independently associated with greater illness severity, we fit a **proportional
odds (ordinal logistic) regression** model using `MASS::polr()`. The outcome
is the six-level ordinal severity scale derived above, where higher levels
indicate more severe illness. An adjusted odds ratio (aOR) greater than 1
indicates that a co-detection group had higher odds of being at an elevated
severity level relative to the reference group (HMPV Only).

Compared to a binary hospitalization outcome, the ordinal approach makes fuller
use of the available clinical information by distinguishing, for example,
between a patient admitted without supplemental oxygen (Hospitalized) and one
admitted to the ICU (ICU). This is particularly relevant for HMPV, which spans
a wide range of severity in this dataset.

The proportional odds model assumes that the effect of each covariate is
constant across all severity thresholds (the *proportional odds assumption*).
P-values are derived from Wald tests; this approximation is asymptotically
valid and appropriate given the dataset size, but should be interpreted with
the usual caveats for sparse upper categories (Ventilated / ECMO and Died).

The model adjusts for **age** (in months, continuous), **preterm birth
status** (dichotomous), **any underlying medical condition** (dichotomous),
and **enrollment site** (categorical). Surveillance season is not adjusted for
in this model; see @tbl-table4 for the season-adjusted sensitivity analysis.

```{r}
#| label: tbl-table2
#| tbl-cap: "Adjusted odds ratios (aOR) for elevated ordinal illness severity by viral co-detection status, from a proportional odds model adjusting for age in months, preterm birth, any underlying medical condition, and enrollment site. An aOR > 1 indicates higher odds of being at a more severe level. The reference group is HMPV Only. Surveillance season is not adjusted for in this model; see Table 4. Only the co-detection term is shown; covariate coefficients are suppressed."

# ── Fit the proportional odds model ───────────────────────────────────────────
#    polr() requires Hess=TRUE to compute standard errors; method="logistic"
#    gives the standard cumulative logit link (equivalent to ordinal logistic)
fit <- polr(c_severity ~ c_agemonths + c_premature + c_xunderlying +
    studysite + codetect_status,
  data   = dat,
  Hess   = TRUE,
  method = "logistic")

# ── Summarize the model ────────────────────────────────────────────────────────
#    exponentiate=TRUE converts log-odds to odds ratios
#    tidy_fun=tidy.polr.wald supplies Wald p-values (see wald_function.R)
#    include=codetect_status suppresses covariate rows from the output
tbl2 <- tbl_regression(
  fit,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()
tbl2
```

Two findings from this model are worth highlighting, each with appropriate caution.

**HMPV/Adeno and higher severity odds.** The HMPV/Adeno group showed higher
adjusted odds of more severe illness. Adenovirus is well-recognized as a cause
of severe respiratory illness in children, including pneumonia and
bronchiolitis.[^2] A true co-infection with active adenovirus disease alongside
HMPV is a plausible driver of greater illness severity, though this analysis
cannot distinguish co-infection from incidental co-detection. These
observations are preliminary and descriptive.

**HMPV/RSV and lower severity odds.** The HMPV/RSV group showed lower adjusted
odds of more severe illness relative to HMPV-alone cases. One possible
explanation is that a meaningful proportion of RSV-positive results in this
group reflect residual shedding of a prior RSV infection rather than true
co-infection with active RSV disease. RSV RNA can remain PCR-detectable for up
to two weeks, and longer in younger or more severely ill children, raising the
possibility that a positive RSV result does not reflect an actively contributing
co-infection.[^3] Notably, HMPV/RSV cases were younger on average than
HMPV-only cases (@tbl-table1), and younger children are typically more likely
to experience severe illness — yet the HMPV/RSV group showed lower severity
odds after adjusting for age. This pattern is consistent with the shedding
hypothesis: if RSV detections in this group were incidental, their clinical
trajectory would reflect HMPV illness alone, potentially explaining why their
severity falls below what their age profile would otherwise predict. The
sensitivity analyses below examine whether underlying conditions or unmeasured
temporal confounding may also contribute to this finding.

[^2]: Lynch JP III, Fishbein M, Echavarria M. Adenovirus. *Semin Respir Crit Care Med*. 2011;32(4):494–511. doi:10.1055/s-0031-1283287.

[^3]: Munywoki PK, Koech DC, Agoti CN, et al. Influence of age, severity of infection, and co-infection on the duration of respiratory syncytial virus (RSV) shedding. *Epidemiol Infect.* 2015;143(4):804–812. doi:10.1017/S0950268814001393.

The stacked bar chart below provides a purely descriptive complement to the
regression results. For each of the three co-detection groups that reached
statistical significance in @tbl-table2 — HMPV Only, HMPV/Adeno, and
HMPV/RSV — we compute the marginal percentage of cases at each ordinal severity
level and display them as filled bars. No model-based adjustment is applied;
the figure reflects the raw observed distribution of severity within each group.
The ordering of groups on the x-axis tracks the direction of the regression
findings: HMPV Only serves as the reference, HMPV/Adeno showed higher severity
odds, and HMPV/RSV showed lower severity odds. Differences visible here are
therefore expected to be directionally consistent with @tbl-table2, though not
adjusted for age, prematurity, underlying conditions, or site.

```{r}
#| label: fig-severity-by-group
#| fig-cap: "Observed distribution of ordinal illness severity among the three co-detection groups that reached statistical significance in the proportional odds model (Table 2). Bars represent the marginal percentage of cases at each severity level within each group. Groups are ordered left to right from reference (HMPV Only) through elevated severity (HMPV/Adeno) to reduced severity (HMPV/RSV). This figure is descriptive and unadjusted."
#| fig-width: 7
#| fig-height: 5
#| echo: false

# ── Subset to the three statistically significant co-detection groups ──────────
SIG.GROUPS <- c("HMPV Only", "HMPV/Adeno", "HMPV/RSV")

sev.counts <- dat[codetect_status %in% SIG.GROUPS,
                  .N, by=.(codetect_status, c_severity)]
sev.counts[, d_perc:=N / sum(N), by=codetect_status]
sev.counts[, codetect_status:=factor(codetect_status, levels=SIG.GROUPS)]

ggplot(sev.counts,
       aes(x=codetect_status, y=d_perc, fill=c_severity)) +
  geom_col(color="black", linewidth=0.3) +
  geom_text(aes(label=sprintf("%.0f%%", 100 * d_perc)),
            position=position_stack(vjust=0.5),
            size=2.8, color="black") +
  scale_fill_brewer(palette="RdYlBu", direction=-1) +
  scale_y_continuous(breaks=seq(0, 1, 0.25),
                     labels=paste0(seq(0, 100, 25), "%"),
                     name="Marginal Percentage") +
  scale_x_discrete(name=NULL) +
  labs(fill="Ordinal Illness Severity") +
  theme(legend.position="bottom")
```

---

# Sensitivity Analysis 1: Patients Without Pre-existing Conditions

In @tbl-table1, the HMPV Only group has a notably higher proportion of patients
with at least one underlying medical condition compared to the HMPV/RSV group.
Although the primary model in @tbl-table2 adjusts for underlying conditions as
a covariate, additive adjustment on the log-odds scale may not fully account
for its effect if the relationship is non-linear or if underlying conditions
are a particularly strong driver of severity in the HMPV-only group. To
evaluate this, we refit the proportional odds model restricting the cohort to
patients **with no pre-existing conditions** (`c_xunderlying == 0`), and
compare results side-by-side with the primary model.

If the HMPV/RSV aOR moves appreciably toward 1.0 in the restricted cohort,
this would suggest that the concentration of underlying conditions in the
HMPV-only group — rather than co-detection status per se — is a primary driver
of the severity difference. If the aOR remains low, the result is more robust
to this potential confound, lending greater relative weight to explanations
such as residual RSV shedding.

Note that `c_xunderlying` is dropped from the right-hand side of the model in
this sensitivity analysis, since the restricted cohort has no variation in that
variable.

```{r}
#| label: tbl-table3
#| tbl-cap: "Sensitivity analysis 1: adjusted odds ratios for elevated ordinal illness severity, restricted to patients with no pre-existing conditions (c_xunderlying == 0). Results are shown side-by-side with the primary model (all patients) for comparison. The reference group is HMPV Only in both models."

# ── Restrict cohort to patients without any underlying conditions ──────────────
dat.nouc <- dat[c_xunderlying==0]

# ── Refit proportional odds model on restricted cohort ────────────────────────
#    c_xunderlying is omitted from the right-hand side (no variation in subset)
fit.nouc <- polr(c_severity ~ c_agemonths + c_premature +
    studysite + codetect_status,
  data   = dat.nouc,
  Hess   = TRUE,
  method = "logistic")

# ── Summarize restricted model ─────────────────────────────────────────────────
tbl3.nouc <- tbl_regression(
  fit.nouc,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()

# ── Side-by-side merge with primary model ─────────────────────────────────────
#    Presents both aORs in a single table for direct comparison; the column
#    spanners label which model each set of columns belongs to
tbl_merge(
  list(tbl2, tbl3.nouc),
  tab_spanner = c("**All Patients**", "**No Pre-existing Conditions**"))
```

**HMPV/Adeno and higher severity odds.** The HMPV/Adeno association with elevated
severity persists in the restricted cohort and strengthens appreciably,
consistent with adenovirus contributing to illness severity independently of
pre-existing conditions. This further supports the hypothesis that HMPV/Adeno
co-detection reflects a true additive viral burden.

**HMPV/RSV and lower severity odds.** The aOR for HMPV/RSV did not attenuate in
the no-underlying-conditions cohort; if anything, it strengthened slightly
relative to the primary model. The imbalance in underlying conditions between
HMPV/RSV and HMPV Only cases therefore does not explain this finding, lending
greater relative weight to the residual RSV shedding hypothesis.

---

# Sensitivity Analysis 2: Season-Adjusted Model

The primary model adjusts for enrollment site but treats observations from
different surveillance seasons as exchangeable. To assess whether unmeasured
temporal confounding — including the substantial disruption in HMPV circulation
during and immediately following the COVID-19 pandemic (seasons 2020–21 and
2021–22) — influences the co-detection findings, we refit the model on the
full cohort with **surveillance season** added as a natural cubic spline with
4 degrees of freedom (`ns(c_ariyear_int, df=4)`). This allows the baseline
severity odds to vary non-linearly across the 11-season study period without
imposing a linear trend assumption.

```{r}
#| label: tbl-table4
#| tbl-cap: "Sensitivity analysis 2: adjusted odds ratios for elevated ordinal illness severity, with surveillance season added as a natural cubic spline (4 df). Results are shown side-by-side with the primary model for comparison. The reference group is HMPV Only in both models."

# ── Refit primary model with season spline on the full cohort ─────────────────
#    ns(c_ariyear_int, df=4) adds 4 basis columns representing a natural cubic
#    spline over the integer season index; knots are placed automatically at
#    the 25th, 50th, and 75th percentiles of c_ariyear_int (seasons 3, 6, 9)
#    with boundary knots at seasons 1 and 11
fit.season <- polr(c_severity ~ c_agemonths + c_premature + c_xunderlying +
    studysite + ns(c_ariyear_int, df=4) + codetect_status,
  data   = dat,
  Hess   = TRUE,
  method = "logistic")

# ── Summarize season-adjusted model ───────────────────────────────────────────
#    The spline term is suppressed by include=codetect_status; its coefficients
#    are not individually interpretable and are not shown
tbl4.season <- tbl_regression(
  fit.season,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()

# ── Side-by-side merge with primary model ─────────────────────────────────────
tbl_merge(
  list(tbl2, tbl4.season),
  tab_spanner = c("**Primary Model**", "**Season-Adjusted**"))
```

Both the HMPV/Adeno and HMPV/RSV effects attenuate after adjusting for season,
with HMPV/Adeno no longer reaching statistical significance. This attenuation
warrants careful interpretation. The season spline absorbs not only the direct
effect of temporal trend on severity, but also any between-season variation in
the prevalence of co-circulating viruses — adenovirus and RSV do not
co-circulate with HMPV at equal rates across seasons, so if co-detection
frequency is itself seasonally structured, the spline may partially absorb the
co-detection signal, producing attenuation that is statistical in nature rather
than biological.

---

# Comparison with Amarin et al. (2026)

This analysis was initially designed to replicate the approach of Amarin et al.
(2026), who examined whether viral co-detections were associated with worse
clinical outcomes among children with laboratory-confirmed RSV. However, the
present analysis now departs from their framework in several important ways:

- **Outcome**: Amarin et al. used binary hospitalization as the primary outcome.
  We use a six-level ordinal severity scale, which captures gradations of
  illness — from routine discharge through ICU admission, mechanical
  ventilation, and death — that a binary outcome would collapse.
- **Study population**: Amarin et al. restricted their cohort to children
  under 2 years of age enrolled from emergency departments or inpatient units.
  The present analysis includes all enrolled ages and care settings.
- **Pathogen**: We study HMPV rather than RSV as the index virus of interest.
- **Sensitivity analyses**: We add (1) a restriction to patients without
  pre-existing conditions and (2) a season-adjusted model using a natural
  cubic spline, neither of which was part of Amarin et al.'s analytical plan.

Because of these differences, results should not be interpreted as a direct
replication and are not numerically comparable to Amarin et al.'s findings.

---

# Additional Considerations

- **RSV shedding and cycle-threshold values**: The interpretation of lower
  severity odds among HMPV/RSV cases as possibly reflecting residual RSV
  shedding could be further explored using CT values. Higher CT values (lower
  viral load) in HMPV/RSV cases relative to HMPV-only cases would be consistent
  with later-stage or co-shedding presentations. CT values for HMPV are
  available for a subset of sites (Houston, Pittsburgh, Rochester, Vanderbilt);
  RSV CT values are pending data request. This comparison is noted as a
  potential extension but is not performed here.

- **Proportional odds assumption**: The proportional odds model assumes that
  the effect of each covariate is constant across all severity thresholds. This
  assumption is not formally tested here and should be evaluated prior to any
  manuscript submission, e.g. using a Brant test (`brant::brant()`).

- **Race/ethnicity**: Two variables are available (`c_race_int` [4 categories]
  and `c_race_int2` [7 categories]). The more granular `c_race_int2` is used
  throughout.

- **Season 11 (2025–26)** was ongoing at the time of the data pull (April 2026).
  Cases from this partial season are included but should be interpreted
  accordingly.

- **Co-detection flag**: The pre-derived variable `c_othercoin6` (any HMPV
  co-detection, research-tested specimens only) is used as the initial flag
  for identifying co-detected cases. Individual pathogen columns are then used
  to assign the specific co-detection group label.

---

*Generated with Quarto and Claude (Anthropic). Contact Ray Pomponio for questions.*
