---
title: "HMPV Co-detections in NVSN: 2015-2026"
subtitle: "Investigating Illness Severity Effects"
author: "Ray Pomponio / Anna Wang-Erickson"
date: today
output-file: "Comprehensive Analysis June 2026.html"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
    df-print: paged
    toc: true
    toc-depth: 3
execute:
  message: false
  warning: false
fig-cap-location: top
---

# Overview

This document examines the association between viral co-detection and illness
severity in children with **human metapneumovirus (HMPV)**, using data from
the New Vaccine Surveillance Network (NVSN) spanning the 2015–16 through
2025–26 surveillance seasons. The analytical framework is informed by Amarin
et al. (2026), who studied co-detection effects in children with RSV, though
the present analysis departs from their approach in several ways described
below.

**Reference:** Amarin JZ, Shaikha AA, Bunsow LI, et al. Respiratory Syncytial
Virus Co-Detection With Other Respiratory Viruses Is Not Significantly
Associated With Worse Clinical Outcomes Among Children Aged <2 Years.
*Open Forum Infectious Diseases*. 2026. <https://doi.org/10.1093/cid/ciaf194>.

The target outputs are:

- **Figure 1**: A flow diagram of study participants, restricted to the
  HMPV-positive analytic cohort.
- **Table 1**: Baseline characteristics and outcomes of HMPV-positive cases,
  stratified by co-detection status (including HCoV and SARS-CoV-2 groups).
- **Figure 2**: An UpSet[^1] plot of the most common co-detection patterns
  among HMPV-positives.
- **Table 2**: Proportional odds regression of adjusted odds ratios for elevated
  illness severity among HMPV-positive cases, with HMPV Only as the reference
  group. HRV, PIV, and influenza co-detections appear significantly associated
  with *lower* severity in the primary model.
- **Figure 3**: Descriptive severity distributions for groups reaching
  statistical significance in Table 2.
- **Sensitivity 1**: Season-adjusted model — HRV, PIV, and influenza effects
  persist, suggesting they are not explained by temporal confounding.
- **Sensitivity 2**: Restricted to patients without pre-existing conditions —
  only HRV survives, raising the possibility that PIV and influenza effects
  are partially driven by case-mix differences.
- **Sensitivity 3**: CT-restricted cohort (CT ≤ 30) approximating confirmed
  active infections — HRV, PIV, and influenza effects disappear entirely,
  strongly suggesting those findings reflect incidental shedding rather than
  true co-infection. HMPV/RSV emerges as significantly associated with
  *elevated* severity, consistent with active RSV co-infection.
- **Sensitivity 4**: RSV-only comparison arm — RSV-alone cases show similar
  severity to HMPV/RSV cases in Sensitivity 3, supporting the interpretation
  that RSV is the dominant pathogen in HMPV/RSV co-detections and HMPV may
  be incidental.
- A placeholder section for infection-order inference using relative CT values.

[^1]: UpSet plots visualize combinations of categorical attributes (here, which viruses were co-detected).

---

# Setup

The following packages are required to run this analysis. Each serves a
distinct purpose:

- **`data.table`**: fast data manipulation for large surveillance datasets
- **`MASS`**: provides `polr()`, which fits the proportional odds (ordinal
  logistic) regression model used in @tbl-table2 through @tbl-table5 and
  in @tbl-table5-rsv
- **`splines`**: provides `ns()` for natural cubic spline basis expansion of
  the season covariate, used in @tbl-table3
- **`consort`**: draws the participant flow diagram
- **`grid`**: low-level graphics support used internally by `consort`
- **`gtsummary`**: produces publication-ready summary tables
- **`gt`**: underlying table engine used by `gtsummary`; required here to
  control column widths in @tbl-table1
- **`ggplot2`** and **`ggpubr`**: general-purpose plotting
- **`ggupset`**: extends `ggplot2` to render UpSet plots

```{r setup}
library(data.table)
library(MASS)
library(splines)
library(consort)
library(grid)
library(gtsummary)
library(gt)
library(ggplot2)
library(ggpubr)
library(ggupset)
source("wald_function.R")
set_consort_defaults(txt_gp=gpar(cex=0.6), parse_markup=TRUE)
theme_set(theme_pubclean(base_size=14))
```

---

# Data Loading and Cleaning

The raw dataset is a CSV file extracted from the NVSN surveillance system.
Before any analysis, several housekeeping steps are required:

1. **Derive hospitalization status** — a simple binary flag (hospitalized vs.
   not) used as a grouping variable in @tbl-table1.
2. **Derive ordinal illness severity** — a six-level ordered outcome ranging
   from routine or ED discharge (Level 1) to death (Level 6). This is the
   primary outcome in @tbl-table2 through @tbl-table5 and @tbl-table5-rsv,
   and captures gradations of severity that a binary hospitalization flag
   would obscure.
3. **Derive minimum CT values per pathogen group** — each pathogen may have
   multiple CT assays (e.g., RSV-untyped, RSV-A, RSV-B). We take the minimum
   CT across all available assays for each pathogen, representing the highest
   observed viral load. This single derived value is used in CT-based
   restrictions and infection-order analyses.
4. **Apply human-readable labels** to coded categorical variables (season,
   site, sex, insurance, race/ethnicity).
5. **Coerce binary (0/1) clinical flags** to integer type.
6. **Backfill missing specimen dates** from the screen date.

```{r load-data}
# ── Load CSV (update path as needed) ──────────────────────────────────────────
cdc <- fread("Data/Pitt_Anna_HMPV_JUN26.csv")

# ── Derive binary hospitalization flag ────────────────────────────────────────
#    c_finalstatus == 1 means the patient was admitted as an inpatient;
#    all other values (ED, outpatient, urgent care) are grouped as "Not Hospitalized"
cdc[, hospital_status:=fcase(
  c_finalstatus==1, "Hospitalized",
  default="Not Hospitalized")]

# ── Derive ordinal illness severity (primary outcome) ─────────────────────────
#    Six levels, assigned in descending priority so each case is classified
#    at its highest observed level of care. Levels 5 and 6 are sparse but
#    retained to avoid information loss at the severe end of the spectrum.
#    Level 1: ED, outpatient, or urgent care discharge
#    Level 2: Admitted to hospital (no supplemental oxygen)
#    Level 3: Admitted + supplemental oxygen (any modality)
#    Level 4: ICU admission
#    Level 5: Mechanical ventilation or ECMO
#    Level 6: Death
cdc[, c_severity := fcase(
  c_died      == 1L,  6L,
  c_intubated == 1L,  5L,
  inptEcmo    == 1L,  5L,
  inptICU     == 1L,  4L,
  hospital_status == "Hospitalized" & (c_suppoxy == 1L | c_blowby == 1L |
    c_hfnc == 1L | c_cpap == 1L),  3L,
  hospital_status == "Hospitalized",  2L,
  default = 1L)]
cdc[, c_severity := factor(c_severity,
  levels  = 1:6,
  labels  = c("Discharge", "Hospitalized", "Hospitalized + O2",
              "ICU", "Ventilated / ECMO", "Died"),
  ordered = TRUE)]

# ── Derive minimum CT per pathogen group ──────────────────────────────────────
#    Where multiple assays exist for the same pathogen (e.g., RSV-untyped,
#    RSV-A, RSV-B), we take the minimum CT across all assays. The minimum CT
#    corresponds to the highest observed viral load for that pathogen group,
#    which is the most conservative representation of active infection.
#    NA values are ignored; a derived CT is NA only if all component assays
#    are NA for that case.
cdc[, d_ct_hmpv  := tmpvCT]
cdc[, d_ct_adeno := tAdenoCT]
cdc[, d_ct_rsv   := pmin(trsvCT, trsvACT, trsvBCT,   na.rm=TRUE)]
cdc[, d_ct_flu   := pmin(tFluApdmH1CT, tFluApdmACT, tFluAH3N2CT,
                         tFluBCT, tFluBvicCT, tFluCCT,
                          na.rm=TRUE)]
cdc[, d_ct_piv   := pmin(tpiv1CT, tpiv2CT, tpiv3CT, tpiv4CT, na.rm=TRUE)]
cdc[, d_ct_rhent := pmin(trhentCT, trhinoCT, tenteroCT, tevd68CT, na.rm=TRUE)]
cdc[, d_ct_hcov  := pmin(tCor229eCT, tCorhku1CT, tCorNL63CT, tCorOC43CT,
                          na.rm=TRUE)]
cdc[, d_ct_scov2 := tsarscov2ctrp]

# ── ARI season labels (11 seasons, 2015-16 through 2025-26) ──────────────────
cdc[, c_ariyear := factor(
  c_ariyear,
  levels = 1:11,
  labels = c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20",
             "2020-21", "2021-22", "2022-23", "2023-24", "2024-25", "2025-26")
)]

# ── Integer season index for spline (1 = 2015-16, 11 = 2025-26) ───────────────
#    ns() requires a numeric covariate; c_ariyear_int is kept separate from
#    the factor c_ariyear which is used for labeling only
cdc[, c_ariyear_int:=as.integer(c_ariyear)]

# ── Site labels ───────────────────────────────────────────────────────────────
#    Site code 7 is not used; Pittsburgh is coded as 8
cdc[, studysite := factor(
  studysite,
  levels = c(1:6, 8),
  labels = c("Vanderbilt", "Rochester", "Cincinnati", "Seattle",
             "Houston", "Kansas City", "Pittsburgh")
)]

# ── Sex labels ────────────────────────────────────────────────────────────────
cdc[, sexch := factor(sexch, levels=c(1, 2), labels=c("Male", "Female"))]

# ── Insurance labels ──────────────────────────────────────────────────────────
#    Code 8 (Unknown) is excluded; unknown cases appear as missing in tables
cdc[, scrinsurance := factor(
  scrinsurance,
  levels = c(1, 2, 3, 4),
  labels = c("Public", "Private", "Both", "None/Self-Pay"))]

# ── Race/ethnicity ────────────────────────────────────────────────────────────
#    Using c_race_int2 (7 categories). NH = Non-Hispanic; AI/AN = American
#    Indian / Alaska Native. Code 8 (Unknown) excluded; appears as missing.
cdc[, c_race_int2 := factor(
  c_race_int2,
  levels = 1:7,
  labels = c("White NH", "Black NH", "Hispanic", "NH Pacific Islander",
             "AI/AN NH", "Asian NH", "Multiple/Other NH"))]

# ── Coerce binary clinical flags to integer 0/1 ───────────────────────────────
#    These variables may import as character strings; explicit coercion avoids
#    errors in table summaries and models.
binary.vars <- c("c_xunderlying", "c_uresp", "c_unuerologic", "c_uoncolimmu",
                 "c_ucardiovasc", "c_premature",
                 "c_respsup", "c_suppoxy", "inptICU", "inptEcmo",
                 "c_blowby", "c_hfnc", "c_cpap", "c_intubated", "c_died",
                 "ANYpneumDx", "ANYbronchDx", "ANYasthDx", "ANYuriDx",
                 "ANYrespfailDx")
cdc[, (binary.vars) := lapply(.SD, function(x) as.integer(as.character(x))),
    .SDcols = binary.vars]
cdc[daycare %in% 0:1, c_daycare:=as.integer(as.character(daycare))]
cdc[smoke %in% 0:1, c_smoke:=as.integer(as.character(smoke))]

# ── Backfill specimen date from screen date where missing ─────────────────────
cdc[is.na(respvisitdt),
    respvisitdt := as.IDate(as.Date(scrdate, format="%m/%d/%Y"))]
```

---

# Inclusion and Exclusion of Participants

Not every enrolled encounter is eligible for this analysis. The flow diagram
below tracks cases from initial enrollment through to the final analytic cohort,
making the reasons for exclusion explicit and auditable.

There are three grounds for exclusion, applied in order:

1. **No research testing** — the specimen was not tested using the research
   multiplex panel, so HMPV status cannot be determined.
2. **HMPV not tested** — the research panel was run, but the HMPV assay was
   not performed (result code 8 = "not performed").
3. **Final status undetermined** — the 10-day chart review needed to classify
   the encounter's highest level of care could not be completed.

After excluding these cases, HMPV-negative and inconclusive results are set
aside (shown as "Undetected" in @fig-consort). Among confirmed HMPV-positive
cases, those with **more than one co-detected virus** are dropped to keep
co-detection groups mutually exclusive and directly comparable. The remaining
cases form the analytic cohort. RSV-positive, HMPV-negative cases are
separately retained for use in Sensitivity Analysis 4 only; they do not
appear in the CONSORT diagram or in any primary analysis.

```{r}
#| label: fig-consort
#| fig-cap: "Flow diagram of HMPV-positive study participants from ARI enrollment through the final analytic cohort. Cases are split at the bottom by co-detection status (HMPV alone vs. HMPV with one co-detected virus). Cases with more than one co-detected virus are dropped to maintain mutually exclusive comparison groups. RSV-positive, HMPV-negative cases are not shown here.."
#| fig-width: 8
#| fig-height: 6
#| echo: false

# ── Define exclusion reasons (applied before HMPV result is known) ────────────
cdc[, exc:=fcase(
  research_tested!=1, "Sample not research-tested",
  tmpv==8 | is.na(tmpv), "Sample not research-tested for HMPV",
  tmpv==0, "HMPV Negative",
  tmpv==2, "HMPV Inconclusive",
  is.na(c_finalstatus), "Final status undetermined")]

# ── Among HMPV-positives, classify co-detection status ────────────────────────
cdc[tmpv==1, codetect:=fcase(
  c_othercoin6==0, "No other virus was detected",
  tAdeno==1, "At least 1 other virus detected",
  c_rsv_result==1, "At least 1 other virus detected",
  anyflu_result==1, "At least 1 other virus detected",
  piv14_pos==1, "At least 1 other virus detected",
  rhent_pos==1, "At least 1 other virus detected",
  c_sarscov2==1, "At least 1 other virus detected",
  hcov_pos==1, "At least 1 other virus detected")]
cdc[is.na(codetect), mis:="Unable to verify codetection status"]

# send QC list to Leah Goldstein (if necessary)
# cdc[tmpv==1 & is.na(codetect), .N]

# ── Count how many co-pathogens were detected (to identify multi-co-detections)
cdc[codetect=="At least 1 other virus detected", codetectn:=rowSums(
  .SD==1, na.rm=TRUE),
  .SDcols=c("tAdeno", "c_rsv_result", "anyflu_result", "piv14_pos",
            "rhent_pos", "c_sarscov2", "hcov_pos")]
cdc[codetectn > 1, more:="More than 1 other virus detected"]

# ── Primary analytic cohort: HMPV-positive, no exclusions, single co-detection
dat <- cdc[is.na(exc) & is.na(mis) & (c_othercoin6==0 | is.na(more))]

# ── RSV-only comparison arm (Sensitivity 4 only) ──────────────────────────────
#    HMPV-negative, RSV-positive, research-tested, same exclusion criteria
dat.rsv <- cdc[!is.na(c_finalstatus) & c_rsv_result==1 & c_hothercoin==0]
dat.rsv[, codetect_status:="RSV Only"]

consort_plot(
  cdc,
  order = c(Caseid    = "Enrolled from outpatient, emergency, or inpatient",
            exc       = "**Excluded**",
            Caseid    = "Human metapneumovirus detected by research testing",
            mis       = "**Missing Data**",
            Caseid    = "HMPV cases with verifiable codetection status",
            codetect  = "",
            more      = "**Dropped**",
            Caseid    = "Analyzed"),
  side_box   = c("exc", "mis", "more"),
  allocation = "codetect")
```

---

# Characteristics of HMPV-Positive Cases

The table below describes the baseline characteristics and clinical outcomes
of all cases in the analytic cohort, stratified by co-detection status.
Each column represents a distinct viral detection group: cases in which HMPV
was the only virus detected, and cases in which HMPV was detected alongside
one specific co-pathogen (adenovirus, influenza, parainfluenza [PIV], RSV,
rhinovirus/enterovirus [HRV], human coronavirus [HCoV], or SARS-CoV-2).

**How to read this table:** For continuous variables (e.g., age in months),
we report the median and interquartile range (IQR). For categorical variables
(e.g., sex, insurance type), we report counts and column percentages. Rows
labeled "Unknown" or missing indicate data that were not collected or could
not be determined. The clinical outcome variables (length of stay, ICU
admission, supplemental oxygen, intubation) appear at the bottom of the table
under *Final Status* and are only meaningful for hospitalized cases;
non-hospitalized cases contribute zeros or missing values to those rows.
The table is paginated (12 rows per page) and horizontally scrollable to
accommodate the six co-detection columns.

```{r}
#| label: tbl-table1
#| tbl-cap: "Baseline demographic characteristics and clinical outcomes of HMPV-positive cases, stratified by co-detection status. Co-detection groups are mutually exclusive (cases with more than one co-detected virus were excluded). Groups include HCoV and SARS-CoV-2 co-detections in addition to the five groups in the primary regression model. Outcome variables (length of stay, ICU admission, supplemental oxygen, and intubation) pertain only to hospitalized cases."

# ── Derive the co-detection status grouping variable ─────────────────────────
#    Priority is assigned in the order listed (Adeno first, HRV last);
#    multi-detection cases have already been excluded, so this is a simple lookup
dat[, codetect_status:=fcase(
  tAdeno==1, "HMPV/Adeno",
  anyflu_result==1, "HMPV/Flu",
  piv14_pos==1, "HMPV/PIV",
  c_rsv_result==1, "HMPV/RSV",
  rhent_pos==1, "HMPV/HRV",
  hcov_pos==1, "HMPV/HCoV",
  c_sarscov2==1, "HMPV/SARS-Cov-2",
  default="HMPV Only")]
dat[, codetect_status:=factor(
  codetect_status,
  levels=c("HMPV Only", "HMPV/Adeno", "HMPV/Flu", "HMPV/PIV",
           "HMPV/RSV", "HMPV/HRV", "HMPV/HCoV", "HMPV/SARS-Cov-2"))]

# ── Variables to include in Table 1, in display order ────────────────────────
tbl1.vars <- c(
  # Demographics
  "c_agemonths", "sexch", "c_race_int2", "scrinsurance", "studysite",
  "c_daycare", "c_smoke",
  # Comorbidities
  "c_xunderlying", "c_uresp", "c_ucardiovasc", "c_unuerologic",
  "c_uoncolimmu", "c_premature",
  # Clinical outcomes (hospital_status first; inpatient-only outcomes follow)
  "hospital_status", "c_los", "inptICU", "c_suppoxy", "c_intubated")

tbl1 <- tbl_summary(
  dat,
  by       = codetect_status,
  include  = all_of(tbl1.vars),
  label    = list(
      c_agemonths     ~ "Age in months, median (IQR)",
      sexch           ~ "Sex",
      c_race_int2     ~ "Race / Ethnicity",
      scrinsurance    ~ "Insurance Type",
      studysite       ~ "Study Site",
      c_daycare       ~ "Day care, preschool, or school attendance",
      c_smoke         ~ "Child lives with someone who smokes tobacco",
      c_xunderlying   ~ "≥1 Underlying Condition",
      c_uresp         ~ "Respiratory Condition",
      c_ucardiovasc   ~ "Cardiovascular Condition",
      c_unuerologic   ~ "Neuromuscular Condition",
      c_uoncolimmu    ~ "Oncologic/Immunosuppressive Condition",
      c_premature     ~ "Prematurity (<37 wks)",
      hospital_status ~ "Final Status",
      c_los           ~ "Length of stay in days, median (IQR)",
      inptICU         ~ "ICU Admission",
      c_suppoxy       ~ "Supplemental Oxygen",
      c_intubated     ~ "Intubation"),
    missing  = "ifany",
    statistic = list(
      all_continuous()  ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)")) |>
  bold_labels() |>
  remove_footnote_header() |>
  modify_header(all_stat_cols() ~ "**{level}**\n**(n={n})**") |>
  as_gt() |>
  cols_width(
    label        ~ px(200),
    everything() ~ px(100)) |>
  opt_interactive(use_compact_mode=TRUE, page_size_default=12)
tbl1
```

---

# Co-detection Patterns

The UpSet plot below visualizes how often HMPV was detected in isolation versus
each possible combination of the other seven respiratory viruses tested in the
research panel (RSV, adenovirus, influenza, parainfluenza, HCoV, SARS-CoV-2, and
rhinovirus/enterovirus). Each bar represents a unique combination of detected
viruses; the dots and connecting lines below the bars indicate which viruses
are included in that combination. Bars are colored by whether the patient was
ultimately hospitalized, providing a first visual impression of whether certain
co-detection patterns are associated with greater illness severity.

Note that this figure is drawn from the **HMPV-positive universe** (before
the restriction to single co-detections), so all combinations — including those
with two or more co-pathogens — are visible here. Only the top 15 most common
combinations are shown.

```{r}
#| label: fig-upset-hmpv
#| fig-cap: "Co-detection patterns among all HMPV-positive cases (research-tested), showing the 15 most frequent combinations of co-detected viruses. Bars are colored by hospitalization status. This figure is drawn before the restriction to single co-detections, so combinations involving multiple co-pathogens are included."
#| echo: false

# ── Named vector mapping display labels to raw column names ───────────────────
lab.paths <- c(
  RSV   = "c_rsv_result",
  Adeno = "tAdeno",
  Flu   = "anyflu_result",
  PIV   = "piv14_pos",
  Rhino = "rhent_pos",
  HCoV  = "hcov_pos",
  `SARS-CoV-2` = "c_sarscov2")

# ── Build a list-column: for each case, which pathogens were detected? ─────────
#    customFun returns the names of detected pathogens (value == 1) per row;
#    stored as a list-column because the number of detections varies by case
customFun <- function(results) names(lab.paths)[results == 1]
cdc[c_othercoin6==1, lst_detections:=apply(.SD, 1, customFun),
    .SDcols=unname(lab.paths)]

ggplot(cdc[codetectn > 0], aes(x=lst_detections, fill=hospital_status)) +
  geom_bar(position=position_dodge(width=1)) +
  scale_x_upset(n_intersections=15) +
  geom_text(stat="count", aes(label=after_stat(count)),
            position=position_dodge(width=1), vjust=-0.5, size=3) +
  scale_y_continuous(name="Number of Cases",
                     expand=expansion(mult=c(0, 0.12))) +
  labs(x="Co-detections with HMPV", fill="Final Status")
```

---

# Illness Severity Odds by Co-detection Status

To assess whether the presence of a viral co-pathogen alongside HMPV is
independently associated with greater illness severity, we fit a **proportional
odds (ordinal logistic) regression** model using `MASS::polr()`. The outcome
is the six-level ordinal severity scale derived above, where higher levels
indicate more severe illness. An adjusted odds ratio (aOR) greater than 1
indicates that a co-detection group had higher odds of being at an elevated
severity level relative to the reference group (HMPV Only).

Compared to a binary hospitalization outcome, the ordinal approach makes fuller
use of the available clinical information by distinguishing, for example,
between a patient admitted without supplemental oxygen (Hospitalized) and one
admitted to the ICU (ICU). This is particularly relevant for HMPV, which spans
a wide range of severity in this dataset.

The proportional odds model assumes that the effect of each covariate is
constant across all severity thresholds (the *proportional odds assumption*).
P-values are derived from Wald tests; this approximation is asymptotically
valid and appropriate given the dataset size, but should be interpreted with
the usual caveats for sparse upper categories (Ventilated / ECMO and Died).

The model adjusts for **age** (in months, continuous), **preterm birth
status** (dichotomous), **any underlying medical condition** (dichotomous),
and **enrollment site** (categorical). Surveillance season is not adjusted for
in this model; see @tbl-table3 for the season-adjusted sensitivity analysis.
HCoV and SARS-CoV-2 co-detections are included as groups in this model;
their sample sizes are small and estimates should be interpreted cautiously.

```{r}
#| label: tbl-table2
#| tbl-cap: "Adjusted odds ratios (aOR) for elevated ordinal illness severity by viral co-detection status, from a proportional odds model adjusting for age in months, preterm birth, any underlying medical condition, and enrollment site. The model is restricted to HMPV-positive cases; the RSV-only comparison arm appears in Sensitivity Analysis 4. An aOR > 1 indicates higher odds of being at a more severe level. The reference group is HMPV Only. Surveillance season is not adjusted for in this model; see Table 3."

# ── Fit the proportional odds model on the HMPV-positive cohort ───────────────
#    polr() requires Hess=TRUE to compute standard errors; method="logistic"
#    gives the standard cumulative logit link (equivalent to ordinal logistic)
fit <- polr(c_severity ~ c_agemonths + c_premature + c_xunderlying +
    studysite + codetect_status,
  data   = dat,
  Hess   = TRUE,
  method = "logistic")

# ── Summarize the model ────────────────────────────────────────────────────────
#    exponentiate=TRUE converts log-odds to odds ratios
#    tidy_fun=tidy.polr.wald supplies Wald p-values (see wald_function.R)
#    include=codetect_status suppresses covariate rows from the output
tbl2 <- tbl_regression(
  fit,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()
tbl2
```

The primary model identifies three co-detection groups with significantly
*lower* adjusted severity odds compared to HMPV alone, each with appropriate caution:

- **HMPV/HRV**, **HMPV/PIV**, and **HMPV/Flu** were each associated with
  reduced odds of more severe illness. Because these are PCR-based detections
  with no CT restriction, a proportion of these positive results may reflect
  residual shedding of a recently resolved infection rather than active
  co-replication at the time of the HMPV illness. If the co-detected virus
  has already cleared clinically, these cases are effectively HMPV-only, which
  would bias their severity toward the HMPV-alone distribution and produce
  the lower aORs observed here. This hypothesis is examined directly in
  Sensitivity Analysis 3 (@tbl-table5).

- **HMPV/RSV** does not reach statistical significance in the primary model.
  This null finding in the unresticted cohort is revisited in Sensitivity
  Analysis 3, where restriction to confirmed active co-infections (CT ≤ 30)
  produces a markedly different result.

These observations are preliminary; the sensitivity analyses below are
essential for distinguishing true co-infection effects from shedding artifacts.

The stacked bar chart below provides a purely descriptive complement to the
primary regression results, showing the raw severity distribution for the
three groups that reached statistical significance in @tbl-table2. No
model-based adjustment is applied. All three groups — HMPV/HRV, HMPV/PIV,
and HMPV/Flu — show a visibly higher proportion of cases at the lowest
severity level (Discharge) compared to HMPV Only, consistent with their
lower aORs. Whether this pattern reflects true biological protection or
shedding misclassification is the central question addressed in
Sensitivity Analysis 3.

```{r}
#| label: fig-severity-by-group
#| fig-cap: "Observed distribution of ordinal illness severity among HMPV-positive co-detection groups reaching statistical significance in the primary proportional odds model (Table 2): HRV, PIV, and influenza. All three show a higher proportion of discharge-level cases relative to HMPV Only, consistent with their lower adjusted severity odds. Whether this reflects true biological protection or incidental shedding of co-detected viruses is examined in Sensitivity Analysis 3. This figure is descriptive and unadjusted."
#| fig-width: 7
#| fig-height: 5
#| echo: false

SIG.GROUPS <- c("HMPV Only", "HMPV/HRV", "HMPV/PIV", "HMPV/Flu")

sev.counts <- dat[codetect_status %in% SIG.GROUPS,
                  .N, by=.(codetect_status, c_severity)]
sev.counts[, d_perc:=N / sum(N), by=codetect_status]
sev.counts[, codetect_status:=factor(codetect_status, levels=SIG.GROUPS)]

ggplot(sev.counts,
       aes(x=codetect_status, y=d_perc, fill=c_severity)) +
  geom_col(color="black", linewidth=0.3) +
  geom_text(aes(label=sprintf("%.0f%%", 100 * d_perc)),
            position=position_stack(vjust=0.5),
            size=2.8, color="black") +
  scale_fill_brewer(palette="RdYlBu", direction=-1) +
  scale_y_continuous(breaks=seq(0, 1, 0.25),
                     labels=paste0(seq(0, 100, 25), "%"),
                     name="Marginal Percentage") +
  scale_x_discrete(name=NULL) +
  labs(fill="Ordinal Illness Severity") +
  theme(legend.position="bottom")
```

---

# Sensitivity Analysis 1: Season-Adjusted Model

The primary model adjusts for enrollment site but treats observations from
different surveillance seasons as exchangeable. To assess whether unmeasured
temporal confounding — including the substantial disruption in HMPV circulation
during and immediately following the COVID-19 pandemic (seasons 2020–21 and
2021–22) — influences the co-detection findings, we refit the model on the
HMPV-positive cohort with **surveillance season** added as a natural cubic
spline with 4 degrees of freedom (`ns(c_ariyear_int, df=4)`). This allows the
baseline severity odds to vary non-linearly across the 11-season study period
without imposing a linear trend assumption.

```{r}
#| label: tbl-table3
#| tbl-cap: "Sensitivity analysis 1: adjusted odds ratios for elevated ordinal illness severity, with surveillance season added as a natural cubic spline (4 df). Results are shown side-by-side with the primary model for comparison. The reference group is HMPV Only in both models. Model is restricted to HMPV-positive cases."

# ── Refit primary model with season spline on the HMPV-positive cohort ─────────
#    ns(c_ariyear_int, df=4) adds 4 basis columns representing a natural cubic
#    spline over the integer season index; knots are placed automatically at
#    the 25th, 50th, and 75th percentiles of c_ariyear_int (seasons 3, 6, 9)
#    with boundary knots at seasons 1 and 11
fit.season <- polr(c_severity ~ c_agemonths + c_premature + c_xunderlying +
    studysite + ns(c_ariyear_int, df=4) + codetect_status,
  data   = dat,
  Hess   = TRUE,
  method = "logistic")

# ── Summarize season-adjusted model ───────────────────────────────────────────
#    The spline term is suppressed by include=codetect_status; its coefficients
#    are not individually interpretable and are not shown
tbl3.season <- tbl_regression(
  fit.season,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()

# ── Side-by-side merge with primary model ─────────────────────────────────────
tbl_merge(
  list(tbl2, tbl3.season),
  tab_spanner = c("**Primary Model**", "**Season-Adjusted**"))
```

All three co-detection groups that were significant in the primary model —
HMPV/HRV, HMPV/PIV, and HMPV/Flu — retain their significantly lower severity
odds after adjusting for season. This rules out temporal confounding as an
explanation for those effects. However, it does not distinguish true
co-infection from shedding: a seasonally distributed shedding artifact
would also survive season adjustment if the shedding prevalence is not
the source of the spurious association. The CT restriction in Sensitivity
Analysis 3 provides a more direct test.

---

# Sensitivity Analysis 2: Patients Without Pre-existing Conditions

The primary model adjusts for any underlying medical condition as a covariate.
However, additive adjustment may not fully capture the effect if the
relationship between comorbidity and severity is non-linear, or if the
distribution of underlying conditions differs substantially across co-detection
groups in ways that confound the lower-severity associations seen for HRV, PIV,
and influenza. To assess this, we refit the model restricting the cohort to
patients **with no pre-existing conditions** (`c_xunderlying == 0`), and
compare results side-by-side with the primary model.

Note that `c_xunderlying` is dropped from the right-hand side of the model in
this sensitivity analysis, since the restricted cohort has no variation in that
variable.

```{r}
#| label: tbl-table4
#| tbl-cap: "Sensitivity analysis 2: adjusted odds ratios for elevated ordinal illness severity, restricted to HMPV-positive patients with no pre-existing conditions (c_xunderlying == 0). Results are shown side-by-side with the primary model (all patients) for comparison. The reference group is HMPV Only in both models."

# ── Restrict HMPV-positive cohort to patients without underlying conditions ────
dat.nouc <- dat[c_xunderlying==0]

# ── Refit proportional odds model on restricted cohort ────────────────────────
#    c_xunderlying is omitted from the right-hand side (no variation in subset)
fit.nouc <- polr(c_severity ~ c_agemonths + c_premature +
    studysite + codetect_status,
  data   = dat.nouc,
  Hess   = TRUE,
  method = "logistic")

# ── Summarize restricted model ─────────────────────────────────────────────────
tbl4.nouc <- tbl_regression(
  fit.nouc,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()

# ── Side-by-side merge with primary model ─────────────────────────────────────
#    Presents both aORs in a single table for direct comparison; the column
#    spanners label which model each set of columns belongs to
tbl_merge(
  list(tbl2, tbl4.nouc),
  tab_spanner = c("**All Patients**", "**No Pre-existing Conditions**"))
```

After restricting to patients without underlying conditions, only HMPV/HRV
retains statistical significance. HMPV/PIV and HMPV/Flu lose significance,
suggesting their lower severity in the primary model is at least partially
explained by the health profile of those groups rather than a direct viral
effect. HMPV/HRV's persistence may reflect a true differential, though it
too disappears after the CT restriction in Sensitivity Analysis 3.

---

# Sensitivity Analysis 3: CT-Restricted Cohort

The analyses above use any positive PCR result to define co-detection status,
regardless of viral load. A positive result at high cycle-threshold (CT) values
may reflect residual viral RNA from a recently resolved infection rather than
active replication. If a substantial proportion of the HRV, PIV, and influenza
co-detections in the primary cohort represent shedding — not active co-infection
— then the lower severity odds observed in @tbl-table2 could be a
misclassification artifact: those cases are clinically HMPV-alone, but are
assigned to a co-detection group, pulling that group's severity toward the
HMPV-only distribution.

To test this, we restrict all detections to cases where the minimum CT value
for each detected pathogen is below **CT ≤ 30**, operationally defining
confirmed active infection for all viruses present. Cases with any detected
pathogen missing a CT value are excluded, since their active/residual status
cannot be determined.

CT values are available for four sites (Houston, Pittsburgh, Rochester, and
Vanderbilt). The restriction is applied consistently across all pathogens for
comparability. The CT ≤ 30 threshold is pragmatic and acknowledged to be
arbitrary; a supplementary analysis at CT ≤ 35 is recommended prior to
manuscript submission.

```{r}
#| label: tbl-table5
#| tbl-cap: "Sensitivity analysis 3: adjusted odds ratios for elevated ordinal illness severity, restricted to HMPV-positive cases from CT-reporting sites (Houston, Pittsburgh, Rochester, Vanderbilt) where all detected pathogens have CT ≤ 30. This approximates a cohort of confirmed active infections. Results are shown side-by-side with the primary model (all sites, no CT restriction) for comparison."

# ── CT threshold constant ─────────────────────────────────────────────────────
CT.THRESHOLD <- 30L

# ── CT-reporting sites ────────────────────────────────────────────────────────
CT.SITES <- c("Houston", "Pittsburgh", "Rochester", "Vanderbilt")

# ── Helper: for a given detection flag and its minimum CT, return TRUE if
#    the case either (a) did not detect the pathogen, or (b) detected it
#    with CT <= threshold. Cases with a positive result but missing CT are
#    excluded (cannot confirm active vs. residual).
active.or.absent <- function(detected, ct, threshold) {
  fifelse(detected==1L,
    !is.na(ct) & ct <= threshold,
    TRUE)
}

# ── Apply CT restriction to HMPV-positive cohort ──────────────────────────────
dat.ct <- dat[
  studysite %in% CT.SITES &
  active.or.absent(tmpv,          d_ct_hmpv,  CT.THRESHOLD) &
  active.or.absent(tAdeno,        d_ct_adeno, CT.THRESHOLD) &
  active.or.absent(anyflu_result, d_ct_flu,   CT.THRESHOLD) &
  active.or.absent(piv14_pos,     d_ct_piv,   CT.THRESHOLD) &
  active.or.absent(c_rsv_result,  d_ct_rsv,   CT.THRESHOLD) &
  active.or.absent(rhent_pos,     d_ct_rhent, CT.THRESHOLD) &
  active.or.absent(hcov_pos,      d_ct_hcov,  CT.THRESHOLD)]

# ── Drop codetect_status levels with fewer than MIN.N cases after CT restriction
MIN.N <- 10L
keep.levels <- dat.ct[, .N, by=codetect_status][N >= MIN.N, codetect_status]
dat.ct <- dat.ct[codetect_status %in% keep.levels]
dat.ct[, codetect_status:=droplevels(codetect_status)]
dat.ct[, studysite:=droplevels(studysite)]

# ── Fit CT-restricted model ───────────────────────────────────────────────────
#    studysite will drop to 4 levels; polr() handles this automatically.
#    Note that the reference level of codetect_status is still "HMPV Only".
fit.ct <- polr(c_severity ~ c_agemonths + c_premature + c_xunderlying +
    studysite + codetect_status,
  data   = dat.ct,
  Hess   = TRUE,
  method = "logistic")

tbl5.ct <- tbl_regression(
  fit.ct,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()

tbl_merge(
  list(tbl2, tbl5.ct),
  tab_spanner = c("**Primary Model (All Sites)**",
                  "**CT ≤ 30 (4 Sites, HMPV-Positive)**"))
```

This sensitivity analysis produces the most substantive findings in the
analysis and directly addresses the shedding hypothesis.

**HRV, PIV, and influenza co-detections lose significance entirely** after
restricting to CT ≤ 30. This pattern is strongly consistent with the
shedding explanation: when only confirmed active co-infections are included,
the apparent lower-severity signal for these groups disappears. The lower
aORs in the primary model were most likely driven by cases where the
co-detected virus had already cleared, making them clinically equivalent to
HMPV-alone presentations. These co-detection findings from the primary model
should therefore not be interpreted as evidence of a biologically meaningful
protective effect.

**HMPV/RSV emerges as significantly associated with *elevated* illness severity**
in the CT-restricted cohort, in contrast to its null finding in the primary
model. This reversal is consistent with the following interpretation: in the
unrestricted cohort, many HMPV/RSV co-detections included cases where RSV
was shedding but no longer clinically active, diluting the severity signal
toward null. When restricted to cases with confirmed active RSV co-infection
(RSV CT ≤ 30 and HMPV CT ≤ 30), the remaining cases appear to involve true
dual active infection — and those cases are more severely ill. This suggests
that active RSV co-infection alongside HMPV carries real clinical significance.
Whether RSV or HMPV is the dominant pathogen in those cases is examined
directly in Sensitivity Analysis 4.

---

# Sensitivity Analysis 4: RSV-Only Comparison Arm

Sensitivity Analysis 3 established that HMPV/RSV cases with confirmed active
co-infection (CT ≤ 30) show elevated illness severity. The natural follow-up
question is whether that elevated severity is attributable to HMPV, to RSV,
or to the combination. If HMPV/RSV cases with active co-infection resemble
RSV-alone cases in their severity distribution, this would strongly suggest
that **RSV is the clinically dominant pathogen** and HMPV is the incidental
detection — the reverse of the study's nominal framing.

To assess this, we expand the cohort to include **RSV-positive, HMPV-negative
cases** as an additional comparison group, fitting the same proportional odds
model with HMPV Only as the reference. The HMPV/RSV aOR in this combined model
can then be compared to the RSV Only aOR to judge their relative severity
profiles.

```{r}
#| label: tbl-table5-rsv
#| tbl-cap: "Sensitivity analysis 4: adjusted odds ratios for elevated ordinal illness severity, with RSV-positive HMPV-negative cases added as a comparison arm. Results are shown side-by-side with the primary HMPV-only model. The reference group is HMPV Only in both models. Coefficient estimates are not directly comparable between models due to the change in the estimation sample; the primary model (Table 2) remains the main analysis."

# ── Build combined dataset with RSV-only arm ──────────────────────────────────
#    dat.rsv was constructed in the CONSORT chunk; codetect_status = "RSV Only"
dat.combined <- rbind(dat, dat.rsv, fill=TRUE)
dat.combined[, codetect_status:=factor(
  codetect_status,
  levels=c("HMPV Only", "HMPV/Adeno", "HMPV/Flu", "HMPV/PIV",
           "HMPV/RSV", "HMPV/HRV", "RSV Only"))]

# ── Fit model on combined cohort ───────────────────────────────────────────────
#    Inclusion of RSV-only cases changes the reference distribution; the key
#    comparison is HMPV/RSV aOR vs RSV Only aOR relative to HMPV Only
fit.rsv <- polr(c_severity ~ c_agemonths + c_premature + c_xunderlying +
    studysite + codetect_status,
  data   = dat.combined,
  Hess   = TRUE,
  method = "logistic")

tbl6.rsv <- tbl_regression(
  fit.rsv,
  tidy_fun     = tidy.polr.wald,
  exponentiate = TRUE,
  include      = codetect_status,
  label        = list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  modify_header(
    label    ~ "**Detection**",
    stat_n   ~ "**N**",
    estimate ~ "**aOR**") |>
  bold_p()

tbl_merge(
  list(tbl2, tbl6.rsv),
  tab_spanner = c("**Primary Model (HMPV-Positive)**",
                  "**With RSV-Only Arm**"))
```

The RSV Only group shows elevated severity odds of similar magnitude to the
HMPV/RSV group observed in Sensitivity Analysis 3. This convergence supports
the conclusion that **RSV is the active driver of illness severity in
HMPV/RSV co-detections**: when RSV is confirmed as actively replicating
(CT ≤ 30), the clinical picture resembles RSV-alone illness. The presence
of HMPV co-detection does not appear to add severity beyond what RSV alone
produces, suggesting HMPV may be incidentally detected in cases where RSV
is the true infecting pathogen.

Taken together with Sensitivity Analysis 3, the data support two conclusions:
(1) co-detection of HRV, PIV, or influenza alongside HMPV does not reflect
true co-infection in most cases — those signals are shedding artifacts; and
(2) when RSV is actively co-detected with HMPV, the illness course is driven
by RSV, not HMPV. These findings are important for the interpretation of
multiplex testing results in clinical practice and motivate further work on
infection-order inference and viral interference.

---

# Placeholder: Infection-Order Inference from Relative CT Values

*This section is reserved for a planned analysis using relative CT values to
approximate the order in which HMPV and RSV infections were acquired.*

The findings in Sensitivity Analyses 3 and 4 suggest that RSV is the
clinically dominant pathogen in HMPV/RSV co-detections. A natural extension
is to ask whether infection order — specifically, whether RSV or HMPV was
acquired first — modifies this relationship. If most HMPV/RSV cases involve
RSV infection followed by HMPV detection (with HMPV possibly representing an
early or resolving infection), this would further support RSV dominance. If
the reverse order is common, it would raise the possibility of viral
interference, whereby an active HMPV infection alters susceptibility to or
the course of a subsequent RSV infection.

The CT difference (RSV CT − HMPV CT) within HMPV/RSV co-detected cases is a
proxy for relative viral loads: a large positive difference (higher RSV CT,
lower RSV load) suggests RSV is declining while HMPV is active; a large
negative difference suggests the opposite. A pre-specified CT difference
threshold will classify cases into infection-order groups; sensitivity across
multiple thresholds (e.g., 3, 5, and 10 CT units) will be examined.

*Implementation is pending discussion with the PI regarding threshold
selection and the grant application framing.*

---

# Comparison with Amarin et al. (2026)

This analysis was initially designed to replicate the approach of Amarin et al.
(2026), who examined whether viral co-detections were associated with worse
clinical outcomes among children with laboratory-confirmed RSV. However, the
present analysis departs from their framework in several important ways and
arrives at substantively different conclusions:

- **Outcome**: Amarin et al. used binary hospitalization as the primary outcome.
  We use a six-level ordinal severity scale, which captures gradations of
  illness — from routine discharge through ICU admission, mechanical
  ventilation, and death — that a binary outcome would collapse.
- **Study population**: Amarin et al. restricted their cohort to children
  under 2 years of age enrolled from emergency departments or inpatient units.
  The present analysis includes all enrolled ages and care settings.
- **Pathogen**: We study HMPV rather than RSV as the index virus of interest,
  and find through CT-based restriction that the RSV/HMPV relationship may be
  better understood with RSV as the dominant pathogen in co-detected cases.
- **Sensitivity analyses**: We add (1) a season-adjusted model, (2) a
  restriction to patients without pre-existing conditions, (3) a CT-restricted
  cohort approximating confirmed active infections, and (4) an RSV-only
  comparison arm. The CT restriction is particularly consequential: it
  reverses the direction of the HMPV/RSV finding and resolves the apparent
  lower-severity signals for HRV, PIV, and influenza as shedding artifacts.

Because of these differences, results should not be interpreted as a direct
replication and are not numerically comparable to Amarin et al.'s findings.

---

# Additional Considerations

- **CT threshold (CT ≤ 30)**: This cutoff is pragmatic and arbitrary. A
  supplementary analysis at CT ≤ 35 is recommended before manuscript
  submission. The biological basis and limitations of the threshold should
  be addressed explicitly in the Methods section of any manuscript.

- **CT site restriction**: The CT-restricted cohort (@tbl-table5) is limited to
  four sites (Houston, Pittsburgh, Rochester, Vanderbilt) and is restricted to
  HMPV-positive cases only. Results should not be treated as representative of
  the full seven-site NVSN network.

- **Minimum CT derivation**: For pathogens with multiple assays, the minimum
  CT across all available assays is used, representing the highest observed
  viral load. NA values are ignored; a derived CT is NA only if all component
  assays are NA.

- **Proportional odds assumption**: The proportional odds model assumes that
  the effect of each covariate is constant across all severity thresholds. This
  assumption is not formally tested here and should be evaluated prior to any
  manuscript submission, e.g. using a Brant test (`brant::brant()`).

- **Race/ethnicity**: Two variables are available (`c_race_int` [4 categories]
  and `c_race_int2` [7 categories]). The more granular `c_race_int2` is used
  throughout.

- **Season 11 (2025–26)** was ongoing at the time of the data pull (June 2026).
  Cases from this partial season are included but should be interpreted
  accordingly.

- **Co-detection flag**: The pre-derived variable `c_othercoin6` (any HMPV
  co-detection, research-tested specimens only) is used as the initial flag
  for identifying co-detected cases. Individual pathogen columns are then used
  to assign the specific co-detection group label.

---

*Generated with Quarto and Claude (Anthropic). Contact Ray Pomponio for questions.*
