---
title: "HMPV Co-detections Preliminary Analysis"
author: "Ray Pomponio"
date: today
output-file: "Preliminary Analysis Mar 2026.html"
format:
  html:
    embed-resources: true
execute:
  echo: false
  warning: false
fig-cap-location: top
---

```{r}
library(data.table)
library(ggplot2)
library(ggpubr)
theme_set(theme_pubclean(base_size=18))
library(gtsummary)
```

```{r}
# load data from CDC; perform superficial clean-up
cdc <- fread("Data/Pitt_Anna_HMPV_DEC25.csv")
dat <- cdc[research_tested==1]
dat[is.na(respvisitdt), respvisitdt:=as.IDate(as.Date(scrdate, format="%m/%d/%Y"))]
dat[tmpv>=2, tmpv:=NA]
dat[c_rsv_result>=2, c_rsv_result:=NA]
```

## Study Population

This analytical dataset contains all NVSN enrollments of acute respiratory illness (ARI), recruited in hospitals, emergency departments, outpatient clinics, and urgent care clinics from `r format(min(dat$respvisitdt), "%b %d %Y")` to `r format(max(dat$respvisitdt), "%b %d %Y")`. Cases were excluded if research testing was not performed (n=`r format(sum(cdc$research_tested!=1), big.mark=",")` exclusions).

## Epidemiological Curves

### Monthly

```{r}
# by months--record no. hmpv, rsv, overall
bmo <- dat[, .(hmpv=sum(tmpv, na.rm=TRUE), rsv=sum(c_rsv_result, na.rm=TRUE), .N),
           by=round(respvisitdt, "months")]
```

```{r}
# by months and age-group--record no. hmpv, rsv, overall
dat[, agegrp:=fcase(c_agemonths < 24, "<2 y.o.", c_agemonths >= 24, "2+ y.o.")]
bma <- dat[, .(hmpv=sum(tmpv, na.rm=TRUE), rsv=sum(c_rsv_result, na.rm=TRUE), .N),
           by=.(round(respvisitdt, "months"), agegrp)]
```

[In @fig-epi-curves-overall we observe seven distinct peaks in RSV case positivity during the period, beginning in the winter of 2017-18. Following these peaks in RSV, we observe lagged peaks in HMPV occurring several months later. Relative timing of the two viruses was consistent after the pandemic. During the pandemic, specifically the winter of 2020-21, we did not observe any positive cases of either virus, possibly due to societal response measures altering the circulation of respiratory viruses.]{.aside}

```{r}
#| label: fig-epi-curves-overall
#| fig-cap: "Epi Curves (case positivity rates) for HMPV and RSV, All ages"
bmo <- melt(bmo, id.vars=c(1, 4), variable.name="virus", value.name="cases")
bmo[, cprt:=cases / N]
ggplot(bmo[order(round, virus)], aes(x=round, y=cprt, fill=virus)) +
  geom_path(aes(color=virus)) +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.15), labels=seq(0, 60, 15)) +
  labs(x="Surveillance month", y="% Positivity")
```

[In @fig-epi-curves-by-age we observe similarity across age groups when identifying peaks in RSV and HMPV case positivity. Patterns of relative timing were generally consistent across age groups, with peaks in RSV occuring several months before peaks in HMPV.]{.aside}

```{r}
#| label: fig-epi-curves-by-age
#| fig-cap: "Epi Curves (case positivity rates) for HMPV and RSV, by Age group"
bma <- melt(bma, id.vars=c(1:2, 5), variable.name="virus", value.name="cases")
bma[, cprt:=cases / N]
ggplot(bma[order(round, virus)], aes(x=round, y=cprt, fill=virus)) +
  facet_wrap(~agegrp, scales="free_y", ncol=1, nrow=2) +
  geom_path(aes(color=virus)) +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.2), labels=seq(0, 60, 20)) +
  labs(x="Surveillance month", y="% Positivity")
```

[In @fig-epi-bars-codetected we observe co-detections represented a minority of overall HMPV detections. However, certain viruses were commonly co-detected with HMPV throughout the period. For example, rhino/enterovirus (orange) was the most commonly co-detected pathogen. Depending on the time period, adenovirus (purple) was co-detected with HMPV in some seasons but not others.]{.aside}

```{r}
#| label: fig-epi-bars-codetected
#| fig-cap: "HMPV-positive Detections by Month, Proportion of co-detections"
dat[!is.na(tmpv), codetected:=fcase(
  tmpv==1 & c_rsv_result==1, "RSV",
  tmpv==1 & anyflu_result==1, "Flu",
  tmpv==1 & piv14_pos==1, "PIV",
  tmpv==1 & tAdeno==1, "Adeno",
  tmpv==1 & rhent_pos==1, "Rhino/Entero",  
  default="None")]
dat[, codetected:=factor(codetected, c("None", "RSV", "Flu", "PIV",
                                       "Adeno", "Rhino/Entero"))]
bmo2 <- dat[!is.na(tmpv), .(hmpv=sum(tmpv, na.rm=TRUE), .N),
           by=.(codetected, round(respvisitdt, "months"))]

ggplot(bmo2, aes(x=round, y=hmpv)) +
  geom_col() +
  geom_col(aes(fill=codetected)) +
  scale_fill_manual(values=c(NA, "red", "blue", "green", "purple",
                             "orange", "brown")) +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  labs(x="Surveillance month", y="HMPV/Co-Detections")
```

### Weekly

```{r}
# weekly percent positivity and co-detection counts
wpp <- dat[between(respvisitdt, as.IDate("2017-01-01"), as.IDate("2024-12-31")),
           .(tested=.N,
             hmpv=sum(tmpv, na.rm=T),
             rsv=sum(c_rsv_result, na.rm=T),
             codetected=sum(tmpv * c_rsv_result, na.rm=T)),
           by=round(as.IDate(respvisitdt), "weeks")]
wpp[, hmpvr:=hmpv / tested]
wpp[, rsvr:=rsv / tested]
```

[In @fig-side-by-side1 I plot the side-by-side epi curves for HMPV and RSV, by week, throughout the study period.]{.aside}

```{r}
#| label: fig-side-by-side1
#| fig-cap: "Epi curves by week."
p1 <- ggplot(wpp[order(round)], aes(x=round, y=hmpvr)) +
  geom_path(color="red") +
  geom_area(alpha=0.25, position="identity") +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.15), labels=seq(0, 60, 15)) +
  labs(x="Surveillance week", y="% Positivity")

p2 <- ggplot(wpp[order(round)], aes(x=round, y=rsvr)) +
  geom_path(color="blue") +
  geom_area(alpha=0.25, position="identity") +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.15), labels=seq(0, 60, 15)) +
  labs(x="Surveillance week", y="% Positivity")

ggarrange(p1, p2, labels=c("HMPV", "RSV"))
```

[In @fig-side-by-side2, I normalized curves to within a moving two-year window. In other words, the maximum value attained is when circulation is highest compared to the preceding or succeeding year. The effect of normalization is to make curves comparable across pathogens, as demonstrated here.]{.aside}

```{r}
#| label: fig-side-by-side2
#| fig-cap: "Epi curves by week, normalized."
for (d in unique(wpp[order(round), round])) {
  dmin <- as.IDate(d) - 365
  dmax <- as.IDate(d) + 365
  wpp[round==d, hmpvn:=hmpvr / wpp[between(round, dmin, dmax), max(hmpvr)]]
  wpp[round==d, rsvn:=rsvr / wpp[between(round, dmin, dmax), max(rsvr)]]
}

p1 <- ggplot(wpp[order(round)], aes(x=round, y=hmpvn)) +
  geom_path(color="red") +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(breaks=seq(0, 1, 0.25), labels=seq(0, 100, 25)) +
  labs(x="Surveillance week", y="Normalized Positivity")

p2 <- ggplot(wpp[order(round)], aes(x=round, y=rsvn)) +
  geom_path(color="blue") +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(breaks=seq(0, 1, 0.25), labels=seq(0, 100, 25)) +
  labs(x="Surveillance week", y="Normalized % Positivity")

ggarrange(p1, p2, labels=c("HMPV", "RSV"))
```

Now for curve-product derivation:

1) Take the product of the normalized curves in fig-side-by-side2.
2) Plot the weekly product throughout the study period; this represents co-circulation patterns.
3) Take the proportion of HMPV detections that were co-detected for the pathogen of interest.
4) Overlay the percentage of co-detections to assess temporal association.

```{r}
#| label: fig-curve-prod-overlay
#| fig-cap: "Overlaid monthly co-detections on weekly co-circulation patterns."
wpp[, prodn:=hmpvn * rsvn]
mpc <- dat[!is.na(tmpv),
    .(hmpv=sum(tmpv, na.rm=TRUE),
      codetected=sum(tmpv * c_rsv_result, na.rm=TRUE)),
    by=.(round(respvisitdt, "months") + 14)]
mpc[, codetectedn:=codetected / hmpv]
ggplot(wpp[order(round)], aes(x=round, y=prodn)) +
  geom_path(color="purple", alpha=0.5) +
  geom_area(alpha=0.25, position="identity") +
  geom_col(data=mpc, aes(x=round, y=codetectedn)) +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(
    breaks=seq(0, 0.5, 0.1), labels=seq(0, 50, 10),
    limits=c(0, max(wpp$prodn, na.rm=TRUE))) +
  labs(x="Surveillance period", y="Product of normalized %s")
```

> I asked if we could then plot the magnitudes of the product curve peaks over time to describe pos vs. negative trend... compare that plot with a separate plot of observed co-detections (% of HMPV cases that are co-detections?) over time to simply ask whether these ines look similar or not.

## Summary Tables

```{r}
#| label: tbl-hmpv-by-pandemic
#| tbl-cap: "HMPV-positive Detections by Pandemic Era"
dat[, era:=fcase(
  respvisitdt <= as.IDate("2020-03-01"), "PRE-pandemic",
  respvisitdt >= as.IDate("2021-01-01"), "POST-pandemic",
  default="PERI-pandemic")]
dat[, era:=factor(era, c("PRE-pandemic", "PERI-pandemic", "POST-pandemic"))]
rvi.cols <- c("tAdeno", "anyflu_result", "rhent_pos", "piv14_pos")
dat[, (rvi.cols) := lapply(.SD, function(x) fcase(x==1, 1, x==0, 0)),
    .SDcols=rvi.cols]
tbl_summary(
  dat[tmpv==1],
  by=era,
  label=list(c_agemonths ~ "Age, months",
             tmpvCT ~ "CT Value, HMPV",
             tAdeno ~ "Adenovirus co-detected",
             c_rsv_result ~ "RSV co-detected",
             anyflu_result ~ "Flu co-detected",
             rhent_pos ~ "Rhino/entero co-detected",
             piv14_pos ~ "PIV co-detected"),
  include=c(c_agemonths, tmpvCT, tAdeno, c_rsv_result, anyflu_result,
            rhent_pos, piv14_pos)) |>
  add_overall(last=TRUE) |>
  modify_header(label="**Surveillance Era**") |>
  modify_footnote_header(
  footnote="Peripandemic was defined as the period between Mar. 1, 2020 and Jan 1. 2021.",
  columns="label")
```


## Findings & Follow-up

### HMPV + RSV

Co-circulation of HMPV and RSV was apparent in all seasons, but most prominent in seasons before the pandemic. It is possibly concerning for an analysis of co-detections, because we will not be able to rule-out the etiology of RSV in determining illness severity. Age group did not appear to modify the co-circulation patterns; therefore, it is reasonable to group all ages together and control for age as a risk factor. Co-circulation was not observed in the 2020-21 season due to the pandemic. We might consider modifying the effect of co-infection by time period, with cases enrolled after 2020 grouped as a temporal category. Such a design would elicit whether effects of co-infection on severity were altered after the pandemic, compared to before the pandemic.

### HMPV + Other Viruses

Co-detection of rhino/enterovirus was relatively common (12% overall), followed by co-detection of adenovirus (5% overall). Most viruses other than RSV were consistently co-detected before versus after the pandemic.

### Next Steps

> Interestingly, she also thought it would be reasonable to take the area under the minimum curve and compare the two methods since I'm not totally sure whether meaningful co-circulation is truly bounded by whichever virus' minimum... it would require establishing thresholds and a Fourier transform to smooth the curve.

> If we wanted to do hypothesis testing... we should look into time series analysis and functional data analysis because the data points within one virus' epi curve are probably auto-correlated.