---
title: "HMPV Co-detections Preliminary Analysis"
author: "Ray Pomponio"
date: today
output-file: "Preliminary Analysis Mar 2026.html"
format:
  html:
    embed-resources: true
    df-print: paged
execute:
  echo: false
  warning: false
fig-cap-location: top
abstract-title: "Summary of findings"
abstract: |
  Co-detections of RSV and HMPV were relatively rare, occurring in <3% of HMPV-positive cases. Co-circulation of RSV and HMPV, as measured by percent positivity in ARI-eligible children, appears to have declined following the pandemic (Mar 1, 2020-Jan 1, 2021), with a proxy of co-circulation peaking at roughly half the magnitude of its pre-pandemic levels. The change in co-circulation patterns may be explained by shifts in the phases of the two Epi curvers, with HMPV percent positivity peaking later relative to RSV in the post-pandemic era. One limitation of this dataset is the inability to pinpoint individual pathogens as the driver of ARI symptoms; while circulation in the entire population may be approximated by ARI  percent positivity, the approach is not without its systematic biases.
---

```{r}
library(data.table)
library(ggplot2)
library(ggpubr)
theme_set(theme_pubclean(base_size=18))
library(gtsummary)
```

```{r}
# load data from CDC; perform superficial clean-up
cdc <- fread("Data/Pitt_Anna_HMPV_DEC25.csv")
cdc[, c_ariyear:=factor(
  c_ariyear,
  c(1:9),
  c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20", "2020-21",
    "2021-22", "2022-23", "2023-24"))]
cdc[, studysite:=factor(
  studysite,
  c(1:6, 8),
  c("Vanderbilt", "Rochester", "Cincinnati",  "Seattle", "Houston",
    "Kansas City", "Pittsburgh"))]
dat <- cdc[research_tested==1]
dat[is.na(respvisitdt), respvisitdt:=as.IDate(as.Date(scrdate, format="%m/%d/%Y"))]
dat[tmpv>=2, tmpv:=NA]
dat[c_rsv_result>=2, c_rsv_result:=NA]
```

## Study Population

This analytical dataset contains all NVSN enrollments of acute respiratory illness (ARI), recruited in hospitals, emergency departments, outpatient clinics, and urgent care clinics from `r format(min(dat$respvisitdt), "%b %d %Y")` to `r format(max(dat$respvisitdt), "%b %d %Y")`. Cases were excluded if research testing was not performed (n=`r format(sum(cdc$research_tested!=1), big.mark=",")` exclusions).

Below is a summary of cases excluded by site and calendar year:

```{r}
exc <- cdc[!Caseid %in% dat$Caseid]
dcast(exc[, .N, by=.(studysite, c_ariyear)], studysite ~ c_ariyear, fill=0)
```


## Swab Source

Of the `r format(nrow(dat), big.mark=",")` cases included in this analysis, `r format(sum(dat$c_resptype==5, na.rm=T), big.mark=",")` were from specimens collected for clinical testing (i.e., "salvaged clinical specimens"). All specimens underwent research laboratory testing, regardless of source type.

## Epidemiological Curves

### Monthly Percent Positivity

```{r}
# by months--record no. hmpv, rsv, overall
bmo <- dat[, .(hmpv=sum(tmpv, na.rm=TRUE), rsv=sum(c_rsv_result, na.rm=TRUE), .N),
           by=round(respvisitdt, "months")]
```

```{r}
# by months and age-group--record no. hmpv, rsv, overall
dat[, agegrp:=fcase(c_agemonths < 24, "<2 y.o.", c_agemonths >= 24, "2+ y.o.")]
bma <- dat[, .(hmpv=sum(tmpv, na.rm=TRUE), rsv=sum(c_rsv_result, na.rm=TRUE), .N),
           by=.(round(respvisitdt, "months"), agegrp)]
```

::: {.panel-tabset}

#### All Ages

[In @fig-epi-curves-overall we observe seven distinct peaks in RSV case positivity during the period, beginning in the winter of 2017-18. Following these peaks in RSV, we observe lagged peaks in HMPV occurring several months later. Relative timing of the two viruses was consistent after the pandemic. During the pandemic, specifically the winter of 2020-21, we did not observe any positive cases of either virus, possibly due to societal response measures altering the circulation of respiratory viruses.]{.aside}

```{r}
#| label: fig-epi-curves-overall
#| fig-cap: "Epi Curves (case positivity rates) for HMPV and RSV, All ages"
bmo <- melt(bmo, id.vars=c(1, 4), variable.name="virus", value.name="cases")
bmo[, cprt:=cases / N]
ggplot(bmo[order(round, virus)], aes(x=round, y=cprt, fill=virus)) +
  geom_path(aes(color=virus)) +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.15), labels=seq(0, 60, 15)) +
  labs(x="Surveillance month", y="% Positivity")
```

#### By Age Group

[In @fig-epi-curves-by-age we observe similarity across age groups when identifying peaks in RSV and HMPV case positivity. Patterns of relative timing were generally consistent across age groups, with peaks in RSV occuring several months before peaks in HMPV.]{.aside}

```{r}
#| label: fig-epi-curves-by-age
#| fig-cap: "Epi Curves (case positivity rates) for HMPV and RSV, by Age group"
bma <- melt(bma, id.vars=c(1:2, 5), variable.name="virus", value.name="cases")
bma[, cprt:=cases / N]
ggplot(bma[order(round, virus)], aes(x=round, y=cprt, fill=virus)) +
  facet_wrap(~agegrp, scales="free_y", ncol=1, nrow=2) +
  geom_path(aes(color=virus)) +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.2), labels=seq(0, 60, 20)) +
  labs(x="Surveillance month", y="% Positivity")
```

:::

### Weekly Percent Positivity

```{r}
# weekly percent positivity and co-detection counts
wpp <- dat[between(respvisitdt, as.IDate("2017-01-01"), as.IDate("2024-12-31")),
           .(tested=.N,
             hmpv=sum(tmpv, na.rm=T),
             rsv=sum(c_rsv_result, na.rm=T),
             codetected=sum(tmpv * c_rsv_result, na.rm=T)),
           by=round(as.IDate(respvisitdt), "weeks")]
wpp[, hmpvp:=hmpv / tested]
wpp[, rsvp:=rsv / tested]
```

::: {.panel-tabset}

#### Unnormalized

```{r}
p1 <- ggplot(wpp[order(round)], aes(x=round, y=hmpvp)) +
  geom_path(color="red") +
  geom_area(alpha=0.25, position="identity") +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.15),
                     labels=seq(0, 60, 15)) +
  labs(x="Surveillance week", y="% Positivity")

p2 <- ggplot(wpp[order(round)], aes(x=round, y=rsvp)) +
  geom_path(color="blue") +
  geom_area(alpha=0.25, position="identity") +
  scale_y_continuous(limits=c(0, 0.6), breaks=seq(0, 0.6, 0.15),
                     labels=seq(0, 60, 15)) +
  labs(x="Surveillance week", y="% Positivity")

ggarrange(p1, p2, labels=c("HMPV", "RSV"))
```

#### Normalized

```{r}
for (d in unique(wpp[order(round), round])) {
  dmin <- as.IDate(d) - 365
  dmax <- as.IDate(d) + 365
  wpp[round==d, hmpvpn:=hmpvp / wpp[between(round, dmin, dmax), max(hmpvp)]]
  wpp[round==d, rsvpn:=rsvp / wpp[between(round, dmin, dmax), max(rsvp)]]
}

p1 <- ggplot(wpp[order(round)], aes(x=round, y=hmpvpn)) +
  geom_path(color="red") +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(breaks=seq(0, 1, 0.25), labels=seq(0, 100, 25)) +
  labs(x="Surveillance week", y="Normalized Positivity")

p2 <- ggplot(wpp[order(round)], aes(x=round, y=rsvpn)) +
  geom_path(color="blue") +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(breaks=seq(0, 1, 0.25), labels=seq(0, 100, 25)) +
  labs(x="Surveillance week", y="Normalized Positivity")

ggarrange(p1, p2, labels=c("HMPV", "RSV"))
```


:::

### Weekly Product of Positivity

::: {.panel-tabset}

#### Unnormalized

```{r}
wpp[, prodp:=hmpvp * rsvp]
ggplot(wpp[order(round)], aes(x=round, y=prodp)) +
  geom_path(color="purple", alpha=0.5) +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(
    breaks=seq(0, 0.05, 0.01), labels=seq(0, 50, 10),
    limits=c(0, max(wpp$prodp, na.rm=TRUE))) +
  labs(x="Surveillance week", y="Product of % Positivity",
       caption="Note: y-axis scale is 10^-3.")
```



#### Normalized

```{r}
wpp[, prodpn:=hmpvpn * rsvpn]
ggplot(wpp[order(round)], aes(x=round, y=prodpn)) +
  geom_path(color="purple", alpha=0.5) +
  geom_area(alpha=0.25, position="identity") +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(
    breaks=seq(0, 0.5, 0.1), labels=seq(0, 50, 10),
    limits=c(0, max(wpp$prodpn, na.rm=TRUE))) +
  labs(x="Surveillance week", y="Product of Norm. Positivity",
       caption="Note: y-axis scale is 10^-2.")
```

:::

### Overlay Monthly Co-detections

[Note, dual axes can be misleading (https://www.datawrapper.de/blog/dualaxis). Co-detections were plotted as a percentage of monthly test results, to account for varying enrollment sizes.]{.aside}

```{r}
mpc <- dat[, .(hmpv=sum(tmpv, na.rm=TRUE), tested=.N,
               codetected=sum(tmpv * c_rsv_result, na.rm=TRUE)),
    by=.(round(respvisitdt, "months") + 14)]
```

::: {.panel-tabset}

#### Unnormalized

```{r}
mpc[, codetectedp:=codetected / tested]
ggplot(wpp[order(round)], aes(x=round, y=prodp)) +
  geom_path(color="purple", alpha=0.5) +
  geom_area(alpha=0.25, position="identity") +
  geom_col(data=mpc, aes(y=codetectedp), alpha=0.75) +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(
    breaks=seq(0, 0.05, 0.01), labels=seq(0, 50, 10),
    limits=c(0, max(wpp$prodp, na.rm=TRUE))) +
  labs(x="Surveillance week", y="Product of % Positivity",
       caption="Note: y-axis scale is 10^-3.")
```


#### Normalized

```{r}
mpc[, codetectedpn:=codetectedp * 10]
ggplot(wpp[order(round)], aes(x=round, y=prodpn)) +
  geom_path(color="purple", alpha=0.5) +
  geom_area(alpha=0.25, position="identity") +
  geom_col(data=mpc, aes(y=codetectedpn), alpha=0.75) +
  scale_x_date(limits=c(as.IDate("2017-01-01"), as.IDate("2024-12-31"))) +
  scale_y_continuous(
    breaks=seq(0, 0.5, 0.1), labels=seq(0, 50, 10),
    limits=c(0, max(wpp$prodpn, na.rm=TRUE))) +
  labs(x="Surveillance week", y="Product of Norm. Positivity",
       caption="Note: y-axis scale is 10^-2.")
```


:::


## Summary Tables

```{r}
#| label: tbl-hmpv-by-pandemic
#| tbl-cap: "HMPV-positive Detections by Pandemic Era"
dat[, era:=fcase(
  respvisitdt <= as.IDate("2020-03-01"), "PRE-pandemic",
  respvisitdt >= as.IDate("2021-01-01"), "POST-pandemic",
  default="PERI-pandemic")]
dat[, era:=factor(era, c("PRE-pandemic", "PERI-pandemic", "POST-pandemic"))]
dat[, codetected:=as.integer(tmpv * c_rsv_result)]
tbl_summary(
  dat,
  by=era,
  label=list(c_agemonths ~ "Age, months",
             tmpv ~ "HMPV detected",
             c_rsv_result ~ "RSV detected",
             codetected ~ "Co-detected"),
  include=c(c_agemonths, tmpv, c_rsv_result, codetected)) |>
  add_overall(last=TRUE) |>
  modify_header(label="**Surveillance Era**") |>
  modify_footnote_header(
  footnote="Peripandemic was defined as the period between Mar. 1, 2020 and Jan 1. 2021.",
  columns="label")
```
