---
title: "HMPV Illness Severity and Co-detections: NVSN 2015–2026"
subtitle: "Replication of Amarin et al. (2026) Framework Applied to HMPV"
author: "Ray Pomponio / Anna Wang-Erickson"
date: today
output-file: "Replication Analysis May 2026.html"
format:
  html:
    embed-resources: true
    df-print: paged
    toc: true
    toc-depth: 3
execute:
  message: false
  warning: false
fig-cap-location: top
---

# Overview

This document replicates, as closely as the NVSN dataset permits, the analytical
approach of Amarin et al. (2026) for RSV, applied here to **human metapneumovirus
(HMPV)** as the pathogen of interest.

 The target outputs are:

- A flow diagram of study participants among ARI-enrolled medical encounters
- **Table 1**: Baseline characteristics and outcomes of HMPV-positive cases,
  stratified by co-detection status 

---

# Setup

```{r setup}
library(data.table)
library(consort)
library(grid)
library(gtsummary)
library(ggplot2)
library(ggpubr)
library(ggupset)

set_consort_defaults(txt_gp=gpar(cex=0.6), parse_markup=TRUE)
theme_set(theme_pubclean(base_size=14))
```


# Data Loading and Cleaning

```{r load-data}
# ── Load CSV (update path as needed) ──────────────────────────────────────────
cdc <- fread("Data/Pitt_Anna_HMPV_APR26.csv")

# ── ARI season labels (11 seasons, 2015-16 through 2025-26) ──────────────────
cdc[, c_ariyear := factor(
  c_ariyear,
  levels = 1:11,
  labels = c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20",
             "2020-21", "2021-22", "2022-23", "2023-24", "2024-25", "2025-26")
)]

# ── Site labels ───────────────────────────────────────────────────────────────
cdc[, studysite := factor(
  studysite,
  levels = c(1:6, 8),
  labels = c("Vanderbilt", "Rochester", "Cincinnati", "Seattle",
             "Houston", "Kansas City", "Pittsburgh")
)]

# ── Sex labels ────────────────────────────────────────────────────────────────
cdc[, sexch := factor(sexch, levels = c(1, 2), labels = c("Male", "Female"))]

# ── Insurance labels ──────────────────────────────────────────────────────────
cdc[, scrinsurance := factor(
  scrinsurance,
  levels = c(1, 2, 3, 4),
  labels = c("Public", "Private", "Both", "None/Self-Pay"))]

# ── Race/ethnicity (use c_race_int2 for finer granularity) ───────────────────
cdc[, c_race_int2 := factor(
  c_race_int2,
  levels = 1:7,
  labels = c("White NH", "Black NH", "Hispanic", "NH Pacific Islander",
             "AI/AN NH", "Asian NH", "Multiple/Other NH"))]

# ── Coerce binary flags to integer 0/1 (some may arrive as character) ─────────
binary.vars <- c("c_xunderlying", "c_uresp", "c_unuerologic", "c_uoncolimmu",
                 "c_ucardiovasc", "c_premature",
                 "c_respsup", "c_suppoxy", "inptICU", "inptEcmo",
                 "c_blowby", "c_hfnc", "c_cpap", "c_intubated", "c_died",
                 "ANYpneumDx", "ANYbronchDx", "ANYasthDx", "ANYuriDx",
                 "ANYrespfailDx")
cdc[, (binary.vars) := lapply(.SD, function(x) as.integer(as.character(x))),
    .SDcols = binary.vars]

# ── Backfill specimen date from screen date where missing ─────────────────────
cdc[is.na(respvisitdt),
    respvisitdt := as.IDate(as.Date(scrdate, format = "%m/%d/%Y"))]
```

---

# Inclusion / Exclusion

```{r}
#| label: fig-consort
#| fig-cap: "Flow diagram of study participants"
#| echo: false

cdc[, exc:=fcase(
  research_tested!=1, "Sample not research-tested",
  tmpv==8, "Sample not research-tested for HMPV",
  is.na(c_finalstatus), "Final status undetermined")]
cdc[, rem:=fcase(
  tmpv==0, "HMPV Negative",
  tmpv==2, "HMPV Inconclusive")]
cdc[, codetect:=fcase(
  c_othercoin6==1, "At least 1 other virus detected",
  default="No other virus was detected")]
cdc[c_othercoin6==1, codetectn:=rowSums(.SD==1),
  .SDcols=c("tAdeno", "anyflu_result", "piv14_pos", "c_rsv_result", "rhent_pos")]
cdc[codetectn > 1, more:="More than 1 other virus detected"]

out <- consort_plot(
  cdc,
  order = c(Caseid    = "Enrolled from outpatient, emergency, or inpatient",
            exc       = "**Excluded**",
            Caseid    = "Tested for human metapneumovirus",
            rem       = "**Undetected**",
            Caseid    = "Human metapneumovirus detected by research testing",
            codetect  = "",
            more      = "",
            Caseid    = "Analyzed"),
  side_box   = c("rem", "exc", "more"),
  allocation = "codetect")
plot(out, grViz = TRUE)

# ── Analytic cohort: singly-detected HMPV cases or single co-detections ───────
dat <- cdc[is.na(exc) & is.na(rem) & (c_othercoin6==0 | is.na(more))]
```

---

# Table 1: Characteristics of HMPV-Positive Cases

Stratified by co-detection status.

```{r}
#| label: tbl-table1
#| tbl-cap: "Demographic and clinical characteristics of HMPV-positive cases by co-detection status"

dat[, codetect_status:=fcase(
  tAdeno==1, "HMPV/Adeno",
  anyflu_result==1, "HMPV/Flu",
  piv14_pos==1, "HMPV/PIV",
  c_rsv_result==1, "HMPV/RSV",
  rhent_pos==1, "HMPV/HRV",
  default="HMPV Only")]
dat[, hospital_status:=fcase(
  c_finalstatus==1, "Hospitalized",
  default="Not Hospitalized")]

tbl1.vars <- c(
  # Demographics
  "c_agemonths", "sexch", "c_race_int2", "scrinsurance", "studysite",
  # Comorbidities
  "c_xunderlying", "c_uresp", "c_ucardiovasc", "c_unuerologic",
  "c_uoncolimmu", "c_premature",
  # Severity
  "hospital_status", "c_los", "inptICU", "c_suppoxy", "c_intubated")

tbl1 <- tbl_summary(
  dat,
  by       = codetect_status,
  include  = all_of(tbl1.vars),
  label    = list(
      c_agemonths     ~ "Age in months, median (IQR)",
      sexch           ~ "Sex",
      c_race_int2     ~ "Race / Ethnicity",
      scrinsurance    ~ "Insurance Type",
      studysite       ~ "Study Site",
      c_xunderlying   ~ "≥1 Underlying Condition",
      c_uresp         ~ "Respiratory Condition",
      c_ucardiovasc   ~ "Cardiovascular Condition",
      c_unuerologic   ~ "Neuromuscular Condition",
      c_uoncolimmu    ~ "Oncologic/Immunosuppressive Condition",
      c_premature     ~ "Prematurity (<37 wks)",
      hospital_status ~ "Final Status",
      c_los           ~ "Length of stay in days, median (IQR)",
      inptICU         ~ "ICU Admission",
      c_suppoxy       ~ "Supplemental Oxygen",
      c_intubated     ~ "Intubation"),
    missing  = "ifany",
    statistic = list(
      all_continuous()  ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)")) |>
  bold_labels() |>
  remove_footnote_header() |>
  modify_caption("Demographic and clinical characteristics of HMPV-positive cases by co-detection status")

tbl1
```


---

# Notes and Caveats

- **Scope**: This document replicates the descriptive and tabular framework of
  Amarin et al. (2026) only. Regression modeling, subgroup analyses, and
  journal-club extensions are intentionally excluded at this stage.

- **c_respsup / c_blowby / c_hfnc / c_cpap** are only available for
  `c_ariyear ≥ 8` (2022-23 onward). For earlier seasons, supplemental-O2
  classification in Severity Level 3 relies on `c_suppoxy` alone. This
  introduces a potential under-ascertainment of Level 3 in earlier seasons;
  consider flagging this in any manuscript methods section.

- **Race/ethnicity**: Two variables are available (`c_race_int` and
  `c_race_int2`). The finer-grained `c_race_int2` is used here. Confirm with
  PI whether this is the preferred variable.

- **HMPV season year 11 (2025-26)** is ongoing as of the data pull date
  (Apr 2026). Consider censoring or labeling partial-season data accordingly.

- Human Coronavirus **(HCoV)** results were withheld from this dataset for unknown
  reasons. **Daycare attendance** and **Smoking exposure** were also withheld.

- **Co-detection flags**: `c_othercoin6` (pre-derived HMPV
  co-detection indicator) used as a flag for all co-detections above.

---

*Generated with Quarto. Contact Ray Pomponio / Anna Wang-Erickson for questions.*
