---
title: "HMPV Illness Severity and Co-detections: NVSN 2015–2026"
subtitle: "Replication of Amarin et al. (2026) Framework Applied to HMPV"
author: "Ray Pomponio / Anna Wang-Erickson"
date: today
output-file: "Replication Analysis May 2026.html"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
    df-print: paged
    toc: true
    toc-depth: 3
execute:
  message: false
  warning: false
fig-cap-location: top
---

# Overview

This document closely follows the analytical approach of Amarin et al. (2026),
who studied the clinical impact of viral co-detections in children with
respiratory syncytial virus (RSV). We apply the same framework here to
**human metapneumovirus (HMPV)** as the pathogen of interest, using data from
the New Vaccine Surveillance Network (NVSN) spanning the 2015–16 through
2025–26 surveillance seasons.

**Reference:** Amarin JZ, Shaikha AA, Bunsow LI, et al. Respiratory Syncytial
Virus Co-Detection With Other Respiratory Viruses Is Not Significantly
Associated With Worse Clinical Outcomes Among Children Aged <2 Years.
*Open Forum Infectious Diseases*. 2026.

The target outputs are:

- A flow diagram of study participants among ARI-enrolled medical encounters
- A table of baseline characteristics and outcomes of HMPV-positive cases,
  stratified by co-detection status
- An UpSet plot of the most common co-detection patterns among HMPV-positives
- A regression summary of adjusted odds ratios of hospital admission status,
  grouped by co-detection status

---

# Setup

The following packages are required to run this analysis. Each serves a
distinct purpose:

- **`data.table`**: fast data manipulation for large surveillance datasets
- **`consort`**: draws the participant flow (CONSORT-style) diagram
- **`grid`**: low-level graphics support used internally by `consort`
- **`gtsummary`**: produces publication-ready summary tables
- **`ggplot2`** and **`ggpubr`**: general-purpose plotting
- **`ggupset`**: extends `ggplot2` to render UpSet plots, which visualize
  combinations of categorical attributes (here, which viruses were co-detected)

```{r setup}
library(data.table)
library(consort)
library(grid)
library(gtsummary)
library(ggplot2)
library(ggpubr)
library(ggupset)

set_consort_defaults(txt_gp=gpar(cex=0.6), parse_markup=TRUE)
theme_set(theme_pubclean(base_size=14))
```

---

# Data Loading and Cleaning

The raw dataset is a CSV file extracted from the NVSN surveillance system.
Before any analysis, several housekeeping steps are required:

1. **Derive hospitalization status** — a simple binary flag (hospitalized vs.
   not) that will serve both as a grouping variable in Table 1 and as the
   outcome in the logistic regression.
2. **Apply human-readable labels** to coded categorical variables (season,
   site, sex, insurance, race/ethnicity), so that all tables and figures are
   self-explanatory without reference to a separate data dictionary.
3. **Coerce binary (0/1) clinical flags** to integer type, correcting for
   the occasional character-type import that would otherwise cause errors
   downstream.
4. **Backfill missing specimen dates** from the screen date, to enable any
   time-series figures that require a valid visit date.

```{r load-data}
# ── Load CSV (update path as needed) ──────────────────────────────────────────
cdc <- fread("Data/Pitt_Anna_HMPV_APR26.csv")

# ── Derive binary hospitalization flag from the "highest level of care" field ─
#    c_finalstatus == 1 means the patient was admitted as an inpatient;
#    all other values (ED, outpatient, urgent care) are grouped as "Not Hospitalized"
cdc[, hospital_status:=fcase(
  c_finalstatus==1, "Hospitalized",
  default="Not Hospitalized")]

# ── ARI season labels (11 seasons, 2015-16 through 2025-26) ──────────────────
#    c_ariyear is stored as an integer 1–11; we convert to a descriptive factor
cdc[, c_ariyear := factor(
  c_ariyear,
  levels = 1:11,
  labels = c("2015-16", "2016-17", "2017-18", "2018-19", "2019-20",
             "2020-21", "2021-22", "2022-23", "2023-24", "2024-25", "2025-26")
)]

# ── Site labels ───────────────────────────────────────────────────────────────
#    studysite is stored as an integer; note that site code 7 is not used,
#    and Pittsburgh is coded as 8
cdc[, studysite := factor(
  studysite,
  levels = c(1:6, 8),
  labels = c("Vanderbilt", "Rochester", "Cincinnati", "Seattle",
             "Houston", "Kansas City", "Pittsburgh")
)]

# ── Sex labels ────────────────────────────────────────────────────────────────
#    Only codes 1 (Male) and 2 (Female) are retained; unknown/missing dropped
cdc[, sexch := factor(sexch, levels = c(1, 2), labels = c("Male", "Female"))]

# ── Insurance labels ──────────────────────────────────────────────────────────
#    Code 8 (Unknown) is excluded here; unknown cases will appear as missing
#    in the summary table
cdc[, scrinsurance := factor(
  scrinsurance,
  levels = c(1, 2, 3, 4),
  labels = c("Public", "Private", "Both", "None/Self-Pay"))]

# ── Race/ethnicity ────────────────────────────────────────────────────────────
#    Using c_race_int2, the more granular of two available race/ethnicity
#    variables (7 categories vs. 4). NH = Non-Hispanic; AI/AN = American
#    Indian / Alaska Native. Code 8 (Unknown) is excluded and will appear
#    as missing in the table.
cdc[, c_race_int2 := factor(
  c_race_int2,
  levels = 1:7,
  labels = c("White NH", "Black NH", "Hispanic", "NH Pacific Islander",
             "AI/AN NH", "Asian NH", "Multiple/Other NH"))]

# ── Coerce binary clinical flags to integer 0/1 ───────────────────────────────
#    These variables represent yes/no clinical events (e.g., ICU admission,
#    intubation). They may import as character strings, so we explicitly
#    convert them to integer to avoid errors in table summaries and models.
binary.vars <- c("c_xunderlying", "c_uresp", "c_unuerologic", "c_uoncolimmu",
                 "c_ucardiovasc", "c_premature",
                 "c_respsup", "c_suppoxy", "inptICU", "inptEcmo",
                 "c_blowby", "c_hfnc", "c_cpap", "c_intubated", "c_died",
                 "ANYpneumDx", "ANYbronchDx", "ANYasthDx", "ANYuriDx",
                 "ANYrespfailDx")
cdc[, (binary.vars) := lapply(.SD, function(x) as.integer(as.character(x))),
    .SDcols = binary.vars]

# ── Backfill specimen date from screen date where missing ─────────────────────
#    respvisitdt is the date the respiratory specimen was collected. Where it
#    is missing, we substitute the screening date (scrdate) as a reasonable
#    approximation of the visit date.
cdc[is.na(respvisitdt),
    respvisitdt := as.IDate(as.Date(scrdate, format = "%m/%d/%Y"))]
```

---

# Inclusion and Exclusion of Participants

Not every enrolled encounter is eligible for this analysis. The flow diagram
below tracks cases from initial enrollment through to the final analytic cohort,
making the reasons for exclusion explicit and auditable.

There are three grounds for exclusion, applied in order:

1. **No research testing** — the specimen was not tested using the research
   multiplex panel, so HMPV status cannot be determined.
2. **HMPV not tested** — the research panel was run, but the HMPV assay was
   not performed (result code 8 = "not performed").
3. **Final status undetermined** — the 10-day chart review needed to classify
   the encounter's highest level of care could not be completed.

After excluding these cases, HMPV-negative and inconclusive results are set
aside (shown as "Undetected" in the diagram). Among confirmed HMPV-positive
cases, those with **more than one co-detected virus** are dropped to keep
co-detection groups mutually exclusive and directly comparable — the same
restriction applied by Amarin et al. (2026). The remaining cases form the
analytic cohort, split by whether any co-pathogen was detected alongside HMPV.

```{r}
#| label: fig-consort
#| fig-cap: "Flow diagram of study participants from ARI enrollment through the final analytic cohort. Cases are split at the bottom by co-detection status (HMPV alone vs. HMPV with one co-detected virus). Cases with more than one co-detected virus are dropped to maintain mutually exclusive comparison groups, following Amarin et al. (2026)."
#| fig-width: 8
#| echo: false

# ── Define exclusion reasons (applied before HMPV result is known) ────────────
cdc[, exc:=fcase(
  research_tested!=1, "Sample not research-tested",
  tmpv==8, "Sample not research-tested for HMPV",
  is.na(c_finalstatus), "Final status undetermined")]

# ── Among tested cases, flag those without a positive HMPV result ─────────────
cdc[, rem:=fcase(
  tmpv==0, "HMPV Negative",
  tmpv==2, "HMPV Inconclusive")]

# ── Among HMPV-positives, classify co-detection status ────────────────────────
cdc[, codetect:=fcase(
  c_othercoin6==1, "At least 1 other virus detected",
  default="No other virus was detected")]

# ── Count how many co-pathogens were detected (to identify multi-co-detections)
cdc[c_othercoin6==1, codetectn:=rowSums(.SD==1),
  .SDcols=c("tAdeno", "c_rsv_result", "anyflu_result", "piv14_pos", "rhent_pos")]
cdc[codetectn > 1, more:="More than 1 other virus detected"]

# ── Final analytic cohort: no exclusions, no ambiguous HMPV result,
#    and at most one co-detected virus ─────────────────────────────────────────
dat <- cdc[is.na(exc) & is.na(rem) & (c_othercoin6==0 | is.na(more))]

consort_plot(
  cdc,
  order = c(Caseid    = "Enrolled from outpatient, emergency, or inpatient",
            exc       = "**Excluded**",
            Caseid    = "Tested for human metapneumovirus",
            rem       = "**Undetected**",
            Caseid    = "Human metapneumovirus detected by research testing",
            codetect  = "",
            more      = "**Dropped**",
            Caseid    = "Analyzed"),
  side_box   = c("rem", "exc", "more"),
  allocation = "codetect")
```

---

# Characteristics of HMPV-Positive Cases

The table below describes the baseline characteristics and clinical outcomes
of all cases in the analytic cohort, stratified by co-detection status.
Each column represents a distinct viral detection group: cases in which HMPV
was the only virus detected, and cases in which HMPV was detected alongside
one specific co-pathogen (adenovirus, influenza, parainfluenza [PIV],
RSV, or rhinovirus/enterovirus [HRV]).

**How to read this table:** For continuous variables (e.g., age in months),
we report the median and interquartile range (IQR) — that is, the middle value
and the range from the 25th to 75th percentile. For categorical variables
(e.g., sex, insurance type), we report counts and column percentages. Rows
labeled "Unknown" or missing indicate data that were not collected or could
not be determined. The clinical outcome variables (length of stay, ICU
admission, supplemental oxygen, intubation) are only meaningful for the
subset of patients who were hospitalized; non-hospitalized cases contribute
zeros or missing values to those rows, and this should be kept in mind when
interpreting those figures.

```{r}
#| label: tbl-table1
#| tbl-cap: "Baseline demographic characteristics and clinical outcomes of HMPV-positive cases, stratified by co-detection status. Co-detection groups are mutually exclusive (cases with more than one co-detected virus were excluded). Outcome variables (length of stay, ICU admission, supplemental oxygen, and intubation) are shown beneath Final Status and pertain only to hospitalized cases."

# ── Derive the co-detection status grouping variable ─────────────────────────
#    Priority is assigned in the order listed (Adeno first, HRV last), which
#    is relevant only for multi-detection cases — but those have already been
#    excluded, so this is effectively a simple lookup.
dat[, codetect_status:=fcase(
  tAdeno==1, "HMPV/Adeno",
  anyflu_result==1, "HMPV/Flu",
  piv14_pos==1, "HMPV/PIV",
  c_rsv_result==1, "HMPV/RSV",
  rhent_pos==1, "HMPV/HRV",
  default="HMPV Only")]

# ── Variables to include in Table 1, in display order ────────────────────────
#    Grouped conceptually: demographics, comorbidities, then clinical outcomes
tbl1.vars <- c(
  # Demographics
  "c_agemonths", "sexch", "c_race_int2", "scrinsurance", "studysite",
  # Comorbidities
  "c_xunderlying", "c_uresp", "c_ucardiovasc", "c_unuerologic",
  "c_uoncolimmu", "c_premature",
  # Clinical outcomes (hospital_status first; inpatient-only outcomes follow)
  "hospital_status", "c_los", "inptICU", "c_suppoxy", "c_intubated")

tbl1 <- tbl_summary(
  dat,
  by       = codetect_status,
  include  = all_of(tbl1.vars),
  label    = list(
      c_agemonths     ~ "Age in months, median (IQR)",
      sexch           ~ "Sex",
      c_race_int2     ~ "Race / Ethnicity",
      scrinsurance    ~ "Insurance Type",
      studysite       ~ "Study Site",
      c_xunderlying   ~ "≥1 Underlying Condition",
      c_uresp         ~ "Respiratory Condition",
      c_ucardiovasc   ~ "Cardiovascular Condition",
      c_unuerologic   ~ "Neuromuscular Condition",
      c_uoncolimmu    ~ "Oncologic/Immunosuppressive Condition",
      c_premature     ~ "Prematurity (<37 wks)",
      hospital_status ~ "Final Status",
      c_los           ~ "Length of stay in days, median (IQR)",
      inptICU         ~ "ICU Admission",
      c_suppoxy       ~ "Supplemental Oxygen",
      c_intubated     ~ "Intubation"),
    missing  = "ifany",
    statistic = list(
      all_continuous()  ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)")) |>
  bold_labels() |>
  remove_footnote_header()

tbl1
```

---

# Co-detection Patterns

The UpSet plot below visualizes how often HMPV was detected alone versus
alongside each possible combination of the other five respiratory viruses
tested in the research panel (RSV, adenovirus, influenza, parainfluenza,
and rhinovirus/enterovirus). Each bar represents a unique combination of
detected viruses; the dots and connecting lines below the bars indicate which
viruses are included in that combination. Bars are colored by whether the
patient was ultimately hospitalized, providing a first visual impression of
whether certain co-detection patterns are associated with greater illness
severity.

Note that this figure is drawn from the **full HMPV-positive universe** (before
the restriction to single co-detections), so all combinations — including those
with two or more co-pathogens — are visible here. Only the top 15 most common
combinations are shown.

```{r}
#| label: fig-upset-hmpv
#| fig-cap: "Co-detection patterns among all HMPV-positive cases (research-tested), showing the 15 most frequent combinations of co-detected viruses. Bars are colored by hospitalization status. This figure is drawn before the restriction to single co-detections, so combinations involving multiple co-pathogens are included."
#| echo: false

# ── Named vector mapping display labels to raw column names ───────────────────
#    This is used to translate the 0/1 columns into a list of pathogen names
#    for each case, which is what the UpSet plot requires.
lab.paths <- c(
  HMPV  = "tmpv",
  RSV   = "c_rsv_result",
  Adeno = "tAdeno",
  Flu   = "anyflu_result",
  PIV   = "piv14_pos",
  Rhino = "rhent_pos")

# ── Build a list-column: for each case, which pathogens were detected? ─────────
#    customFun receives a row of 0/1 values and returns the names of those
#    that equal 1 (i.e., were detected). The result is stored as a list-column
#    because the number of detected pathogens varies by case.
customFun <- function(results) names(lab.paths)[results == 1]
cdc[, lst_detections:=apply(.SD, 1, customFun), .SDcols=unname(lab.paths)]

ggplot(cdc[tmpv==1], aes(x=lst_detections, fill=hospital_status)) +
  geom_bar(position=position_dodge(width=1)) +
  scale_x_upset(n_intersections=15) +
  geom_text(stat="count", aes(label=after_stat(count)),
            position=position_dodge(width=1), vjust=-0.5, size=3) +
  scale_y_continuous(name="Number of Cases",
                     expand=expansion(mult=c(0, 0.12))) +
  labs(x="Co-detections with HMPV", fill="Final Status")
```

---

# Admission Odds by Co-detection Status

To assess whether the presence of a viral co-pathogen alongside HMPV is
independently associated with a higher likelihood of hospitalization, we
fit a logistic regression model. The outcome is whether the patient was
hospitalized (vs. discharged from the ED, outpatient clinic, or urgent care).

The model adjusts for potential confounders that are known to influence
illness severity and hospitalization risk in children with acute respiratory
illness: **age** (in months, continuous), **preterm birth status**
(dichotomous), **any underlying medical condition** (dichotomous), and
**enrollment site** (categorical, to account for systematic differences in
care practices across the seven NVSN hospitals). After adjusting for these
factors, the remaining association between co-detection group and
hospitalization is expressed as an **adjusted odds ratio (aOR)**.

An aOR greater than 1 indicates that a particular co-detection group had
higher odds of hospitalization compared to the reference group (HMPV Only),
after accounting for age, prematurity, underlying conditions, and site.
Confidence intervals and p-values are provided; results in bold meet the
conventional p < 0.05 threshold for statistical significance.

```{r}
#| label: tbl-table2
#| tbl-cap: "Adjusted odds ratios (aOR) of hospitalization by viral co-detection status, from a logistic regression model adjusting for age in months, preterm birth, any underlying medical condition, and enrollment site. The reference group is HMPV Only. Only the co-detection term is shown; covariate coefficients are suppressed."

# ── Fit the logistic regression model ─────────────────────────────────────────
#    The outcome is a binary indicator of hospitalization derived from
#    hospital_status. Covariates are entered on the right-hand side;
#    codetect_status is the exposure of interest (reference = "HMPV Only").
fit <- glm(hospital_status=="Hospitalized" ~ c_agemonths + c_premature +
             c_xunderlying + studysite + codetect_status,
           family=binomial,
           data=dat)

# ── Summarize the model ────────────────────────────────────────────────────────
#    exponentiate=TRUE converts log-odds coefficients to odds ratios.
#    include=c(codetect_status) suppresses the display of covariate rows,
#    showing only the co-detection term that is the focus of inference.
#    add_n and add_nevent append the number of cases and hospitalized events
#    per group, to aid interpretation.
tbl2 <- tbl_regression(
  fit,
  exponentiate=TRUE,
  include=c(codetect_status),
  intercept=FALSE,
  label=list(codetect_status ~ NA)) |>
  add_n(location="level") |>
  add_nevent(location="level") |>
  modify_header(label ~ "**Detection**", stat_nevent ~ "**Admitted**") |>
  bold_p()
tbl2
```

---

# Notes and Caveats

## Comparison with Amarin et al. (2026)

This analysis replicates the design of Amarin et al. (2026), who examined
whether viral co-detections were associated with worse clinical outcomes among
children with laboratory-confirmed RSV. Their key methodological choices —
restricting to single co-detections, adjusting for age and underlying
conditions, and using hospitalization as the primary outcome — are preserved here.

**The most important difference** between this analysis and Amarin et al. is
the study population. Amarin et al. restricted their cohort to **children
under 2 years of age**, a group with well-established vulnerability to RSV
and the primary target of RSV prevention strategies. **The present analysis
includes all enrolled ages** in the NVSN dataset and does not apply that
age restriction. This means our results are not directly comparable to
Amarin et al. on the basis of the outcome alone, and any interpretation of
co-detection effects should be considered in the context of a broader and
more heterogeneous age distribution.

## Additional Caveats

- **Respiratory support variables** (`c_respsup`, `c_blowby`, `c_hfnc`,
  `c_cpap`) are only available for seasons 2022–23 onward (`c_ariyear ≥ 8`).
  For earlier seasons, supplemental oxygen classification relies on `c_suppoxy`
  alone, which may lead to under-ascertainment of oxygen support in those years.

- **Race/ethnicity**: Two variables are available (`c_race_int` [4 categories]
  and `c_race_int2` [7 categories]). The more granular `c_race_int2` is used
  throughout.

- **Season 11 (2025–26)** was ongoing at the time of the data pull (April 2026).
  Cases from this partial season are included but should be interpreted
  accordingly; seasonal-level summaries for 2025–26 are incomplete.

- **Human Coronavirus (HCoV)** results were withheld from this dataset for
  unknown reasons. Data on **daycare attendance** and **smoking exposure**,
  which Amarin et al. included as covariates, were also withheld and could
  not be incorporated.

- **Co-detection flag**: The pre-derived variable `c_othercoin6` (any HMPV
  co-detection, research-tested specimens only) is used as the initial flag for
  identifying co-detected cases. Individual pathogen columns are then used to
  assign the specific co-detection group label.

---

*Generated with Quarto. Contact Ray Pomponio (raymond.pomponio@outlook.com) for questions.*
